\documentclass[12pt]{article}
\usepackage{amsmath, amssymb, enumitem, geometry, xcolor}
\geometry{margin=1in}

\title{Lesson 22: Practice Problems - Finding Potential Functions}
\author{ODE 1 - Prof. Adi Ditkowski}
\date{}

\usepackage{mdframed}
\usepackage{tikz}
\begin{document}
\maketitle

\section*{Part A: Basic Potential Functions (Problems 1-6)}

\begin{enumerate}
\item Find the potential function: $(2x + 3y)dx + (3x + 4y)dy = 0$

\item Find the potential function: $(y^{2} + 2xy)dx + (2xy + x^{2})dy = 0$

\item Find the potential function: $(e$^{x}$\sin y + 2x)dx + (e$^{x}$\cos y)dy = 0$

\item Find the potential function: $(3x^{2} + 6xy^{2})dx + (6x$^{2y}$ + 4y^{3})dy = 0$

\item Find the potential function: $\left(\frac{1}{x} + y\right)dx + \left(x - \frac{1}{y}\right)dy = 0$

\item Find the potential function: $(2xy^{3} + 1)dx + (3x$^{2y}$^2 + 2)dy = 0$
\end{enumerate}

\section*{Part B: Method Comparison (Problems 7-11)}

\begin{enumerate}[start=7]
\item Solve using both Method 1 and Method 2, verify they agree:
$(y\cos x + 2x)dx + (\sin x + 2y)dy = 0$

\item Solve using Method 1, then verify using Method 2:
$(2xe$^{y}$ + y)dx + (x$^{2e}$^y + x + 3y^{2})dy = 0$

\item Solve using Method 2, then verify using Method 1:
$(y$^{2e}$^{xy} + 2x)dx + (2xy$e^{xy}$ + 3y^{2})dy = 0$

\item Use Method 3 (line integral from origin):
$(2xy + y^{2})dx + (x^{2} + 2xy)dy = 0$

\item Compare all three methods for:
$(3x$^{2y}$ + y^{3})dx + (x^{3} + 3xy^{2})dy = 0$
\end{enumerate}

\section*{Part C: Initial Value Problems (Problems 12-16)}

\begin{enumerate}[start=12]
\item Solve $(2xy + 3)dx + (x^{2} - 1)dy = 0$ with $y(1) = 2$

\item Solve $(y$e^{xy}$ + 2x)dx + (x$e^{xy}$ + 2y)dy = 0$ with $y(0) = 1$

\item Solve $\left(\cos y - y\sin x\right)dx + \left(-x\sin y - \cos x\right)dy = 0$ with $y(\pi/2) = \pi$

\item Solve $(2x + y^{2})dx + (2xy - 3y^{2})dy = 0$ with $y(2) = 1$

\item Find the solution curve passing through $(1, 1)$ for:
$(3x$^{2y}$^2 + 2x)dx + (2x$^{3y}$ + 3y^{2})dy = 0$
\end{enumerate}

\section*{Part D: Complex Expressions (Problems 17-21)}

\begin{enumerate}[start=17]
\item Find $H$ for: $\left(e$^{x}$ + \frac{y}{x^{2}}\right)dx + \left(\frac{1}{x} + e$^{y}$\right)dy = 0$

\item Find $H$ for: $(y\sec^{2} x + \sec x \tan x)dx + (\tan x + 2y)dy = 0$

\item Find $H$ for: $\left(\frac{2x}{x^{2} + y^{2}} + 3x^{2}\right)dx + \left(\frac{2y}{x^{2} + y^{2}} + 2y\right)dy = 0$

\item Find $H$ for: $(\ln y + 2xy)dx + \left(\frac{x}{y} + x^{2} + y^{2}\right)dy = 0$

\item Find $H$ for: $(y^{2}\cos x - 3x$^{2y}$)dx + (2y\sin x - x^{3} + \ln y)dy = 0$
\end{enumerate}

\section*{Part E: Verification Problems (Problems 22-25)}

\begin{enumerate}[start=22]
\item Given $H(x,y) = x$^{3y}$ + xy^{3} + x^{2}$, find $M$ and $N$ such that $Mdx + Ndy = 0$ is exact with this potential function.

\item A student claims $H = x$^{2y}$ + y^{3}$ is the potential function for $(2xy)dx + (x^{2} + 3y^{2})dy = 0$. Verify or correct this.

\item Find the error: For $(2xy + 1)dx + (x^{2} + 2y)dy = 0$, a student got $H = x$^{2y}$ + x + y^{2} + 5$. Is this correct?

\item Show that if $H_{1}$ and $H_{2}$ are both potential functions for the same exact equation, then $H_{1} - H_{2} = $ constant.
\end{enumerate}

\section*{Part F: Exam-Style Problems (Problems 26-30)}

\begin{enumerate}[start=26]
\item (Prof. Ditkowski 2023) Find the potential function using two methods:
$(2xy + y^{2}\cos x)dx + (x^{2} + 2y\sin x)dy = 0$

\item Given that $(ax$^{2y}$ + 2xy^{2})dx + (x^{3} + bx$^{2y}$)dy = 0$ is exact with $a = 3$, find $b$ and solve the equation.

\item The equation $(P(x) + y^{2})dx + (Q(y) + 2xy)dy = 0$ is exact.
\begin{enumerate}[label=(\alph*)]
    \item Find the relationship between $P(x)$ and $Q(y)$
    \item If $P(x) = x^{2}$, find $Q(y)$ and the potential function
\end{enumerate}

\item Find all functions $f(x)$ such that $(f(x) + y)dx + (x + e$^{y}$)dy = 0$ is exact, then solve for $f(x) = x$.

\item A potential function satisfies $H(0,y) = y^{2}$ and $H(x,0) = x^{3}$. If the equation is $(3x^{2} + g(y))dx + (h(x) + 2y)dy = 0$, find $g(y)$, $h(x)$, and $H(x,y)$.
\end{enumerate}

\section*{Solutions and Key Insights}

\textbf{Problem 1:} Using Method 1:
$H = \int(2x + 3y)dx = x^{2} + 3xy + g(y)$
$\frac{\partial H}{\partial y} = 3x + g'(y) = 3x + 4y$
So $g'(y) = 4y$, giving $g(y) = 2y^{2}$
Answer: $H = x^{2} + 3xy + 2y^{2}$

\textbf{Problem 7:}
Method 1: $H = \int(y\cos x + 2x)dx = y\sin x + x^{2} + g(y)$
$\frac{\partial H}{\partial y} = \sin x + g'(y) = \sin x + 2y$
So $g(y) = y^{2}$. Thus $H = y\sin x + x^{2} + y^{2}$

Method 2: $H = \int(\sin x + 2y)dy = y\sin x + y^{2} + f(x)$
$\frac{\partial H}{\partial x} = y\cos x + f'(x) = y\cos x + 2x$
So $f(x) = x^{2}$. Same answer!

\textbf{Problem 12:} First find $H$:
$H = \int(2xy + 3)dx = x$^{2y}$ + 3x + g(y)$
$\frac{\partial H}{\partial y} = x^{2} + g'(y) = x^{2} - 1$
So $g(y) = -y$, giving $H = x$^{2y}$ + 3x - y$

Using $y(1) = 2$: $C = (1)^{2}(2) + 3(1) - 2 = 3$
Solution: $x$^{2y}$ + 3x - y = 3$

\textbf{Problem 22:} Given $H = x$^{3y}$ + xy^{3} + x^{2}$
$M = \frac{\partial H}{\partial x} = 3x$^{2y}$ + y^{3} + 2x$
$N = \frac{\partial H}{\partial y} = x^{3} + 3xy^{2}$

\textbf{Key Strategy Note:} For Problem 19, use the fact that $d(\ln(x^{2} + y^{2})) = \frac{2x\,dx + 2y\,dy}{x^{2} + y^{2}}$

\textbf{Warning for Problem 30:} Use the boundary conditions to build $H$ systematically.

\end{document}