\documentclass[12pt]{article}
\usepackage{amsmath, amssymb, amsthm, enumitem, geometry}
\usepackage{mdframed, xcolor}
\geometry{margin=1in}

\newmdenv[linecolor=blue,linewidth=1pt]{hint}
\newcommand{\solution}[1]{\textcolor{blue}{\textbf{Solution Hint:} #1}}

\title{Lesson 40: Practice Problems\\
\large Critical Points and Linearization}
\author{ODE 1 - Prof. Adi Ditkowski}
\date{}

\usepackage{tikz}
\begin{document}
\maketitle

\section*{Part A: Finding Critical Points (5 problems)}

\begin{enumerate}
\item Find all critical points of the system:
$$\dot{x} = x - y, \quad \dot{y} = x^2 - 4$$
\solution{Set both equal to zero: $y = x$ and $x^2 = 4$, giving two critical points.}

\item Find all critical points of:
$$\dot{x} = y(1-x^2), \quad \dot{y} = x - y$$
\solution{Consider cases: either $y = 0$ or $x^2 = 1$. Check all combinations.}

\item Determine the critical points of:
$$\dot{x} = \sin(x), \quad \dot{y} = y - \cos(x)$$
\solution{From $\sin(x) = 0$, we get $x = n\pi$. Then find corresponding $y$ values.}

\item Find all equilibria of the predator-prey system:
$$\dot{x} = x(2 - x - y), \quad \dot{y} = y(x - 1)$$
\solution{Factor first, then consider cases systematically.}

\item For what values of $a$ does the system have exactly one critical point?
$$\dot{x} = ax - y, \quad \dot{y} = x^2 + y^2 - 1$$
\solution{Analyze when the system $ax - y = 0$ and $x^2 + y^2 = 1$ has unique solution.}
\end{enumerate}

\section*{Part B: Jacobian and Linearization (6 problems)}

\begin{enumerate}[resume]
\item Compute the Jacobian matrix for:
$$\dot{x} = x^2 - y + 1, \quad \dot{y} = xy$$
\solution{$J = \begin{pmatrix} 2x & -1 \\ y & x \end{pmatrix}$}

\item Find the Jacobian at the origin for:
$$\dot{x} = \sin(x) + y, \quad \dot{y} = e^x - e^y$$
\solution{Remember: $\frac{d}{dx}\sin(x)|_{x=0} = \cos(0) = 1$ and $\frac{d}{dx}e^x|_{x=0} = 1$}

\item Linearize the pendulum equation at $(0, 0)$:
$$\dot{x} = y, \quad \dot{y} = -\sin(x) - 0.5y$$
\solution{Use $\sin(x) \approx x$ near origin.}

\item Find the linearization of the following system at $(1, 1)$:
$$\dot{x} = y - x^3, \quad \dot{y} = -x + y^3$$
\solution{First verify $(1,1)$ is a critical point, then compute $J(1,1)$.}

\item Compute the Jacobian for the Lotka-Volterra system:
$$\dot{x} = x(\alpha - \beta y), \quad \dot{y} = y(-\gamma + \delta x)$$
at the coexistence equilibrium.
\solution{First find the equilibrium: $x = \gamma/\delta$, $y = \alpha/\beta$.}

\item For the system $\dot{x} = y + x^2$, $\dot{y} = -x + y^2$, find all critical points and compute the Jacobian at each.
\solution{From first equation: $y = -x^2$. Substitute into second.}
\end{enumerate}

\section*{Part C: Stability Classification (5 problems)}

\begin{enumerate}[resume]
\item Classify the critical point $(0,0)$ for:
$$\dot{x} = -2x + y, \quad \dot{y} = x - 3y$$
\solution{Find eigenvalues of $J = \begin{pmatrix} -2 & 1 \\ 1 & -3 \end{pmatrix}$}

\item Determine the stability of all critical points:
$$\dot{x} = y, \quad \dot{y} = x - x^3$$
\solution{Three critical points: $(0,0)$, $(1,0)$, $(-1,0)$. Classify each.}

\item For what values of $\mu$ is the origin stable?
$$\dot{x} = -x + \mu y, \quad \dot{y} = -\mu x - y$$
\solution{Eigenvalues depend on $\mu$. Find stability condition.}

\item Classify the critical point $(1, 0)$ for:
$$\dot{x} = x(1-x) - xy, \quad \dot{y} = y(x-2)$$
\solution{Compute $J(1,0)$ carefully.}

\item Determine the type and stability of the origin for:
$$\dot{x} = y + x^3, \quad \dot{y} = -2x - 3y + xy$$
\solution{Nonlinear terms vanish at origin, so linearization is exact there.}
\end{enumerate}

\section*{Part D: Hartman-Grobman Applications (4 problems)}

\begin{enumerate}[resume]
\item For which critical points does linearization determine the local behavior?
$$\dot{x} = y, \quad \dot{y} = \sin(x) - y$$
\solution{Check if eigenvalues have non-zero real parts (hyperbolic condition).}

\item Consider the system $\dot{x} = -y + x^2$, $\dot{y} = x + y^2$.
At which critical point(s) might linearization fail to determine stability?
\solution{Find where eigenvalues are purely imaginary.}

\item For the system:
$$\dot{x} = y^2 - x, \quad \dot{y} = x^2 - y$$
Identify all hyperbolic equilibria.
\solution{Check each critical point for zero real parts in eigenvalues.}

\item Given that linearization at $(0,0)$ yields eigenvalues $\lambda = \pm i$, what can you conclude about the nonlinear system's behavior near the origin?
\solution{Linearization inconclusive; could be center, spiral, or more complex.}
\end{enumerate}

\section*{Part E: Exam-Style Problems (5 problems)}

\begin{enumerate}[resume]
\item \textbf{[Prof. Ditkowski Style]} Consider the competing species model:
$$\dot{x} = x(3 - x - 2y), \quad \dot{y} = y(2 - x - y)$$
\begin{enumerate}[label=(\alph*)]
    \item Find ALL critical points
    \item Compute the Jacobian at each critical point
    \item Classify each equilibrium and state its stability
    \item Which equilibria are biologically meaningful?
\end{enumerate}
\solution{Four critical points expected. Check axes and interior.}

\item \textbf{[Comprehensive]} For the system:
$$\dot{x} = y - x^2, \quad \dot{y} = -x - y + 2$$
\begin{enumerate}[label=(\alph*)]
    \item Show that $(1, 1)$ is a critical point
    \item Find any other critical points
    \item Linearize at each critical point
    \item Sketch the expected behavior near each equilibrium
\end{enumerate}
\solution{Quadratic in $x$ means at most 2 critical points.}

\item \textbf{[Bifurcation Preview]} Consider:
$$\dot{x} = \mu x - y - x^3, \quad \dot{y} = x + \mu y - y^3$$
\begin{enumerate}[label=(\alph*)]
    \item Show $(0,0)$ is always a critical point
    \item Find the linearization at the origin
    \item For what values of $\mu$ is the origin stable?
    \item What happens at $\mu = 0$?
\end{enumerate}
\solution{Eigenvalues are $\lambda = \mu \pm i$.}

\item \textbf{[Physical System]} A nonlinear oscillator satisfies:
$$\ddot{x} + \dot{x} - x + x^3 = 0$$
\begin{enumerate}[label=(\alph*)]
    \item Rewrite as a first-order system
    \item Find all equilibria
    \item Determine stability of each equilibrium
    \item Interpret physically
\end{enumerate}
\solution{Let $y = \dot{x}$ to get system form.}

\item \textbf{[Mixed Techniques]} For the gradient system:
$$\dot{x} = -\frac{\partial V}{\partial x}, \quad \dot{y} = -\frac{\partial V}{\partial y}$$
where $V(x,y) = x^4 + y^4 - 2x^2 - 2y^2 + xy$:
\begin{enumerate}[label=(\alph*)]
    \item Find all critical points
    \item Classify each using linearization
    \item What can you say about stability using $V$ directly?
\end{enumerate}
\solution{Critical points occur where $\nabla V = 0$. Use second derivative test.}
\end{enumerate}

\section*{Part F: Additional Practice (5 problems)}

\begin{enumerate}[resume]
\item Find and classify all equilibria of:
$$\dot{x} = x - x^3, \quad \dot{y} = -y$$

\item For the Hamiltonian system with $H(x,y) = \frac{1}{2}y^2 + \frac{1}{2}x^2 - \frac{1}{3}x^3$:
$$\dot{x} = \frac{\partial H}{\partial y}, \quad \dot{y} = -\frac{\partial H}{\partial x}$$
Find and classify all critical points.

\item Consider the chemical reaction system:
$$\dot{x} = 1 - xy^2, \quad \dot{y} = xy^2 - y$$
Find the unique positive equilibrium and determine its stability.

\item For what values of $a$ and $b$ is the origin an asymptotically stable spiral?
$$\dot{x} = ax + y, \quad \dot{y} = -x + by$$

\item \textbf{[Challenge]} Consider the system:
$$\dot{x} = y + x(1 - x^2 - y^2), \quad \dot{y} = -x + y(1 - x^2 - y^2)$$
\begin{enumerate}[label=(\alph*)]
    \item Show that the origin is the only critical point inside $x^2 + y^2 < 1$
    \item Classify the origin using linearization
    \item Show that $x^2 + y^2 = 1$ is an invariant circle
    \item Describe the global behavior
\end{enumerate}
\end{enumerate}

\begin{hint}
\textbf{Key Strategies for Prof. Ditkowski's Exam:}
\begin{itemize}
    \item Always find ALL critical points - check systematically
    \item Show Jacobian computation explicitly - no shortcuts
    \item State both type AND stability clearly
    \item When eigenvalues have zero real part, state "linearization inconclusive"
    \item For transcendental equations, graphical analysis is acceptable
    \item Double-check arithmetic - especially signs in the Jacobian
\end{itemize}
\end{hint}

\end{document}