\documentclass[12pt]{article}
\usepackage{amsmath, amssymb, enumitem}
\usepackage{geometry}
\geometry{margin=1in}

\title{Practice Problems: Implicit and Singular Solutions}
\author{Lesson 13 Exercises}

\begin{document}
\maketitle

\section*{Part A: Implicit Solutions - Recognition and Verification (6 problems)}

\begin{enumerate}
\item Verify that $x^2 - xy + y^2 = C$ is an implicit solution of some ODE. Find the ODE.

\item Show that $e^y + e^{-x} = C$ solves $\frac{dy}{dx} = e^{y-x}$

\item For the circles $x^2 + y^2 = r^2$, find the differential equation they satisfy.

\item Verify that $\sin(x + y) = x$ defines an implicit solution. Find the corresponding ODE.

\item Can $xy + \ln(xy) = C$ be solved explicitly for $y$? Find the ODE it satisfies.

\item Show that the family of ellipses $\frac{x^2}{a^2} + \frac{y^2}{b^2} = 1$ (with $a/b = k$ constant) satisfies a certain ODE.
\end{enumerate}

\section*{Part B: Finding Singular Solutions - Basic (6 problems)}

\begin{enumerate}[resume]
\item Find all solutions of $(y')^2 = 4y$, including any singular solutions.

\item For the general solution $y = Cx + C^2$, find the singular solution using the C-discriminant.

\item Show that $y = 0$ is a singular solution of $y' = 2\sqrt{|y|}$

\item Find the singular solution of $(y')^2 - xy' + y = 0$ (Clairaut type).

\item Given general solution $(y - Cx)^2 = 1 + C^2$, find the envelope.

\item Determine if $y(y')^2 - 2xy' + 4x = 0$ has singular solutions.
\end{enumerate}

\section*{Part C: Clairaut's Equation and Variants (6 problems)}

\begin{enumerate}[resume]
\item Solve completely: $y = xy' + \sqrt{1 + (y')^2}$

\item Find general and singular solutions: $y = xy' - \frac{1}{y'}$

\item Solve: $y = xy' + a(y')^2$ where $a$ is a constant.

\item Transform and solve: $2y = xy' + \ln(y')$

\item Find all solutions: $y - xy' = e^{y'}$

\item Solve the generalized Clairaut: $y = x f(y') + g(y')$
\end{enumerate}

\section*{Part D: p-Discriminant Method (5 problems)}

\begin{enumerate}[resume]
\item Use p-discriminant to find singular solutions of: $y = x(y')^2 - (y')^3$

\item Apply p-discriminant to: $(y')^3 - xy' + y = 0$

\item Find singular solutions: $x(y')^2 - 2yy' + 4x = 0$

\item Determine all solutions of: $8(y')^3 = 27y^2$ using p-discriminant.

\item Show the p-discriminant process for: $y^2(1 + (y')^2) = 1$
\end{enumerate}

\section*{Part E: Parametric Solutions (5 problems)}

\begin{enumerate}[resume]
\item Express solutions parametrically: $(y')^2 + 2y' = x$

\item Find parametric solution: $x = (y')^3 + y'$

\item Solve using parameter $p = y'$: $y = 2xy' + (y')^3$

\item Convert to parametric form: $(1 + (y')^2)^3 = (y')^2$

\item Find parametric solution: $x^2 + (y')^2 = 1$
\end{enumerate}

\section*{Part F: Advanced and Mixed Problems (6 problems)}

\begin{enumerate}[resume]
\item Show that the family $y^2 = 2Cx + C^2$ has envelope $y^2 + 2x = 0$. Verify both satisfy the same ODE.

\item For $(y')^2(1 - x^2) = (1 - y^2)$:
    \begin{enumerate}
        \item Find implicit general solution
        \item Identify any singular solutions
        \item Sketch solution curves
    \end{enumerate}

\item The brachistochrone problem gives $(1 + (y')^2) = \frac{C}{y}$. Find parametric solution using $\theta$ where $y' = \cot(\theta/2)$.

\item Consider $y^2 = (x + c)^3$:
    \begin{enumerate}
        \item Find the differential equation
        \item Show $y = 0$ is singular
        \item Find the C-discriminant
    \end{enumerate}

\item For the Lagrange equation $y = xf(y') + g(y')$ where $f(p) \neq p$:
    \begin{enumerate}
        \item Derive the general solution method
        \item Apply to $y = 2xy' + (y')^2$
    \end{enumerate}

\item Prove that if $F(x,y,C) = 0$ is the general solution and its envelope exists, the envelope satisfies the same ODE.
\end{enumerate}

\section*{Part G: Exam-Style Problems (7 problems)}

\begin{enumerate}[resume]
\item \textbf{[Prof. Ditkowski Special]} Consider $(y')^2 = (1 - y^2)$
    \begin{enumerate}
        \item Solve by separation of variables
        \item Find all singular solutions
        \item Show that $|y| \leq 1$ for all solutions
        \item Sketch the complete solution family
        \item Identify the envelope
    \end{enumerate}

\item \textbf{[Clairaut Comprehensive]} For $y = xy' - \frac{(y')^3}{3}$:
    \begin{enumerate}
        \item Find the general solution
        \item Derive the singular solution
        \item Show the singular solution is an envelope
        \item Find points where general and singular solutions meet
        \item Sketch several solution curves
    \end{enumerate}

\item \textbf{[Implicit Challenge]} The equation $x^3 + y^3 = 3axy$ (folium of Descartes):
    \begin{enumerate}
        \item Find the ODE it satisfies
        \item Show solution exists implicitly near any non-singular point
        \item Find where vertical tangents occur
        \item Determine singular points
    \end{enumerate}

\item \textbf{[C-Discriminant Application]} Given $(y - cx)^2 = c^3$:
    \begin{enumerate}
        \item Apply C-discriminant method
        \item Verify the discriminant satisfies the ODE
        \item Interpret geometrically
        \item Find where uniqueness fails
    \end{enumerate}

\item \textbf{[Verification Focus]} A student claims $y^2 = x$ is a singular solution of $2(y')^3 = 3y'x - 3y$:
    \begin{enumerate}
        \item Verify it satisfies the ODE
        \item Find the general solution
        \item Prove it's singular
        \item Find the relationship between solutions
    \end{enumerate}

\item \textbf{[Parametric Excellence]} For $(y')^2 + (y')^3 = x$:
    \begin{enumerate}
        \item Find parametric solution with $p = y'$
        \item Determine where solution has vertical tangent
        \item Find range of $x$ values
        \item Express arc length parametrically
    \end{enumerate}

\item \textbf{[Prof. Ditkowski Comprehensive]}
    Consider Riccati-type: $y' = y^2 - x^2$
    \begin{enumerate}
        \item Show no elementary explicit solution exists
        \item Find solution satisfying $y(0) = 0$ as power series (first 4 terms)
        \item Determine if singular solutions exist
        \item Find curves where $y' = 0$
        \item Analyze behavior as $|x| \to \infty$
    \end{enumerate}
\end{enumerate}

\section*{Answer Key with Essential Hints}

\textbf{1.} ODE: $\frac{dy}{dx} = \frac{2x - y}{x - 2y}$

\textbf{7.} General: $y = (x + C)^2$; Singular: $y = 0$

\textbf{8.} Singular solution: $y = -x^2/4$ (parabola)

\textbf{10.} Singular: $x^2 + 4y = 0$ (use $p = x/2$)

\textbf{13.} General: $y = Cx + \sqrt{1 + C^2}$; Singular: $x^2 + y^2 = 1$

\textbf{16.} $y - xy' = e^{y'}$ not exactly Clairaut; use substitution

\textbf{19.} Set $y' = p$: $y = xp^2 - p^3$; eliminate using $\partial/\partial p = 0$

\textbf{24.} Parametric: $x = p^2 + 2p$, $y = \frac{2p^3}{3} + p^2 + C$

\textbf{31.} C-discriminant gives $x = -C/2$, substitute back

\textbf{37.} General: $y = \sin(x + C)$ or $y = -\sin(x + C)$; Singular: $y = \pm 1$

\textbf{38.} General: $y = Cx - C^3/3$; Singular: $y = \frac{2x^{3/2}}{3\sqrt{3}}$ for $x \geq 0$

\textbf{41.} Apply C-discriminant: $27y^2 = 4x^3$ (semicubical parabola)

\textbf{43.} Power series: $y = -\frac{x^3}{3} + \frac{2x^7}{63} + ...$

\end{document}