\documentclass[12pt]{article}
\usepackage{amsmath, amssymb, enumitem, geometry}
\geometry{margin=1in}

\title{Lesson 29: Practice Problems}
\author{Liouville's Formula and Applications}
\date{}

\usepackage{mdframed}
\usepackage{xcolor}
\usepackage{tikz}
\begin{document}
\maketitle

\section*{Part A: Basic Liouville Calculations (6 problems)}

\begin{enumerate}
\item Use Liouville's formula to find $W(t)$ for $\mathbf{x}' = \begin{bmatrix} 3 & 1 \\ 0 & -2 \end{bmatrix}\mathbf{x}$ with $W(0) = 5$.

\item Calculate the Wronskian at $t = 2$ for the system $\mathbf{x}' = \begin{bmatrix} 1 & 0 & 0 \\ 0 & 2 & 0 \\ 0 & 0 & -3 \end{bmatrix}\mathbf{x}$ if $W(0) = 1$.

\item Find $W(t)$ for solutions of $y''' - 3y'' + 2y' - y = 0$ with $W(0) = 2$.

\item Given $\mathbf{x}' = \begin{bmatrix} \cos t & \sin t \\ -\sin t & \cos t \end{bmatrix}\mathbf{x}$, find $W(t)$ using Liouville.

\item For the harmonic oscillator $\mathbf{x}' = \begin{bmatrix} 0 & 1 \\ -\omega^{2} & 0 \end{bmatrix}\mathbf{x}$, verify that volume is preserved.

\item Calculate $W(\pi)$ for $\mathbf{x}' = \begin{bmatrix} \sin t & 0 \\ 0 & \cos t \end{bmatrix}\mathbf{x}$ with $W(0) = 3$.
\end{enumerate}

\section*{Part B: Trace and Eigenvalue Connections (5 problems)}

\begin{enumerate}[resume]
\item A $3 \times 3$ system has eigenvalues $\lambda_{1} = 2$, $\lambda_{2} = -1$, $\lambda_{3} = -4$. Find $W(t)/W(0)$.

\item If $\text{tr}(A) = -5$ for a constant matrix, and $W(1) = $e^{-5$}$, find $W(3)$.

\item The characteristic polynomial is $\lambda^{3} - 2\lambda^{2} - 5\lambda + 6 = 0$. Find $\text{tr}(A)$ and describe $W(t)$ behavior.

\item Given eigenvalues $1 \pm 2i$, find the trace and Wronskian evolution for the $2 \times 2$ system.

\item A system has $W(t) = 3$e^{-6t$}$. If two eigenvalues are $-1$ and $-2$, find the third.
\end{enumerate}

\section*{Part C: Stability Analysis (5 problems)}

\begin{enumerate}[resume]
\item Determine stability of $\mathbf{x}' = \begin{bmatrix} -2 & 1 \\ 3 & -4 \end{bmatrix}\mathbf{x}$ using Liouville's formula.

\item For what values of $a$ is the system $\mathbf{x}' = \begin{bmatrix} a & 1 \\ -1 & -a \end{bmatrix}\mathbf{x}$ volume-preserving?

\item Analyze stability of $y'' + 3y' + 2y = 0$ using the trace of its companion matrix.

\item Given $\ddot{x} + b\dot{x} + 4x = 0$, find $b$ values for which the Wronskian decays.

\item Determine long-term behavior of $W(t)$ for $\mathbf{x}' = \begin{bmatrix} t & 1 \\ 0 & -t \end{bmatrix}\mathbf{x}$.
\end{enumerate}

\section*{Part D: Special Systems (5 problems)}

\begin{enumerate}[resume]
\item Show that the system $\mathbf{x}' = \begin{bmatrix} 0 & 1 & 0 \\ -1 & 0 & 1 \\ 0 & -1 & 0 \end{bmatrix}\mathbf{x}$ preserves volume.

\item For the Hamiltonian system with $H = \frac{1}{2}(p^{2} + q^{2})$, verify Liouville's theorem.

\item Find all $2 \times 2$ matrices $A$ with $\text{tr}(A) = 0$ and $\det(A) = 1$.

\item Prove that skew-symmetric matrices ($A$^{T}$ = -A$) always have trace zero.

\item For the periodic system $\mathbf{x}' = \begin{bmatrix} \cos(2t) & 0 \\ 0 & -\cos(2t) \end{bmatrix}\mathbf{x}$, find $W(2\pi)/W(0)$.
\end{enumerate}

\section*{Part E: Applications and Theory (4 problems)}

\begin{enumerate}[resume]
\item Use Liouville to prove that if all eigenvalues have negative real parts, then $W(t) \to 0$ as $t \to \infty$.

\item Show that for the equation $y^{(n)} + a_{n-1}y^{(n-1)} + \cdots + a$_{0y}$ = 0$, the Wronskian satisfies $W(t) = W(0)$e^{-a_{n-1}t$}$.

\item If two solutions have Wronskian $W_{12}(t) = $e^{3t$}$, what can you conclude about the trace of the system matrix?

\item Prove that similar matrices have the same trace, hence the same Wronskian evolution.
\end{enumerate}

\section*{Part F: Exam-Style Problems (5 problems)}

\begin{enumerate}[resume]
\item (Prof. Ditkowski style) Consider $\mathbf{x}' = \begin{bmatrix} 1 & 2 & 0 \\ 0 & -1 & 3 \\ 0 & 0 & 2 \end{bmatrix}\mathbf{x}$:
\begin{enumerate}[label=(\alph*)]
\item Find $\text{tr}(A)$ and all eigenvalues
\item Use Liouville to find $W(t)$ with $W(0) = 1$
\item Verify using direct eigenvalue sum
\item Determine stability of the origin
\item Find $\lim_{t \to \infty} W(t)$
\end{enumerate}

\item The damped oscillator $\ddot{x} + 2\zeta\omega\dot{x} + \omega$^{2x}$ = 0$ has parameters $\omega = 2$, $\zeta = 0.5$.
\begin{enumerate}[label=(\alph*)]
\item Write as a first-order system
\item Find the trace
\item Calculate $W(t)/W(0)$
\item How long until the Wronskian decreases by factor of $e$?
\end{enumerate}

\item Given that solutions to a third-order system have $W(t) = 5$e^{-3t$}$:
\begin{enumerate}[label=(\alph*)]
\item Find $\text{tr}(A)$
\item If $\lambda_{1} = 2$ is an eigenvalue, and the other two are complex conjugates, find them
\item Write a possible matrix $A$
\item Analyze stability
\end{enumerate}

\item For the time-dependent system $\mathbf{x}' = \begin{bmatrix} e$^{t}$ & 0 \\ 0 & -e$^{t}$ \end{bmatrix}\mathbf{x}$:
\begin{enumerate}[label=(\alph*)]
\item Compute $\text{tr}(A(t))$
\item Find $W(t)$ using Liouville with $W(0) = 2$
\item Determine if volume is preserved
\item Find $\lim_{t \to \infty} W(t)$
\end{enumerate}

\item (Comprehensive) Consider the fourth-order equation $y^{(4)} - 2y''' - 3y'' + 4y' + 4y = 0$.
\begin{enumerate}[label=(\alph*)]
\item Convert to a system and find $\text{tr}(A)$
\item Use Liouville to express $W(t)$
\item Given that $\lambda_{1} = 2$ and $\lambda_{2} = -1$ are eigenvalues, find the others
\item Verify $\sum \lambda$_{i}$ = \text{tr}(A)$
\item Determine the long-term behavior of solutions
\item Is the zero solution stable?
\end{enumerate}
\end{enumerate}

\section*{Solutions and Hints}

\textbf{Selected Solutions:}

\textbf{Problem 1:} $\text{tr}(A) = 3 + (-2) = 1$, so $W(t) = 5e$^{t}$$

\textbf{Problem 2:} $\text{tr}(A) = 1 + 2 + (-3) = 0$, so $W(2) = W(0) = 1$

\textbf{Problem 5:} $\text{tr}(A) = 0 + 0 = 0$, volume preserved!

\textbf{Problem 7:} $W(t) = W(0) \cdot $e^{(2-1-4)t}$ = W(0) \cdot $e^{-3t$}$

\textbf{Problem 12:} $\text{tr}(A) = -2 + (-4) = -6 < 0$, system is stable

\textbf{Problem 13:} Need $a + (-a) = 0$ for all $a$, so always volume-preserving

\textbf{Problem 17:} $\text{tr}(A) = 0 + 0 + 0 = 0$, confirms volume preservation

\textbf{Problem 22:} Average trace over period is zero, so $W(2\pi) = W(0)$

\textbf{Problem 26:} $\text{tr}(A) = 1 + (-1) + 2 = 2$, so $W(t) = $e^{2t$}$, unstable

\textbf{Problem 29:} $\text{tr}(A) = -3$, so $\lambda_{2} + \lambda_{3} = -5$. With conjugates $a \pm bi$: $2a = -5$, so $\lambda_{2,3} = -2.5 \pm bi$

\textbf{Key Formulas:}
\begin{itemize}
\item Liouville: $W(t) = W(t_{0})$e^{\int_{t_{0}$}$^{t}$ \text{tr}(A(s))ds}$
\item Constant case: $W(t) = W(0)$e^{\text{tr}(A)$ \cdot t}$
\item Trace-eigenvalue: $\text{tr}(A) = \sum \lambda$_{i}$$
\item Scalar $n$th-order: trace = $-a_{n-1}$
\item Volume preserved $\Leftrightarrow \text{tr}(A) = 0$
\end{itemize}

\textbf{Stability Quick Check:}
\begin{itemize}
\item $\text{tr}(A) < 0 \Rightarrow W(t) \to 0$ (stable tendency)
\item $\text{tr}(A) > 0 \Rightarrow W(t) \to \infty$ (unstable)
\item $\text{tr}(A) = 0 \Rightarrow W(t) = $ constant (neutral)
\end{itemize}

\end{document}