\documentclass[12pt]{article}
\usepackage{amsmath, amssymb, amsthm, enumitem, geometry}
\geometry{margin=1in}

\title{Lesson 32: Practice Problems - Complex Eigenvalues}
\author{ODE 1 - Prof. Adi Ditkowski}
\date{}

\usepackage{mdframed}
\usepackage{xcolor}
\usepackage{tikz}
\begin{document}
\maketitle

\section*{Part A: Finding Complex Eigenvalues and Eigenvectors}

\begin{enumerate}
\item Find the eigenvalues of $A = \begin{pmatrix} 0 & -1 \\ 1 & 0 \end{pmatrix}$ and classify the equilibrium.

\item Find the eigenvalues and eigenvectors of $A = \begin{pmatrix} 2 & -5 \\ 1 & 2 \end{pmatrix}$.

\item Determine $\alpha$ and $\beta$ if the eigenvalues are $\lambda = 3 \pm 4i$.

\item For what values of $a$ does $A = \begin{pmatrix} a & -2 \\ 2 & a \end{pmatrix}$ have complex eigenvalues?

\item Find a 2\times2 matrix with eigenvalues $\lambda = -1 \pm 3i$.
\end{enumerate}

\section*{Part B: Extracting Real Solutions}

\begin{enumerate}[start=6]
\item Given eigenvalue $\lambda = 2i$ with eigenvector $\mathbf{v} = \begin{pmatrix} 1 \\ -i \end{pmatrix}$, find two real solutions.

\item Extract real solutions from $\lambda = 1 + i$ with eigenvector $\mathbf{v} = \begin{pmatrix} 1 \\ -i \end{pmatrix}$.

\item If $\mathbf{x}$_{c}$(t) = $e^{(2+3i)t}\begin{pmatrix}$ 1+i \\ 2 \end{pmatrix}$, find the real and imaginary parts.

\item Convert the complex solution $\mathbf{x}(t) = $e^{-t+2it}\begin{pmatrix}$ i \\ 1 \end{pmatrix}$ to real form.

\item Given $\lambda = 3 - 2i$ and $\mathbf{v} = \begin{pmatrix} 2 \\ 1+i \end{pmatrix}$, find the general real solution.
\end{enumerate}

\section*{Part C: 2\times2 Systems with Complex Eigenvalues}

\begin{enumerate}[start=11]
\item Solve: $\mathbf{x}' = \begin{pmatrix} 0 & -4 \\ 1 & 0 \end{pmatrix}\mathbf{x}$

\item Solve the IVP: $\mathbf{x}' = \begin{pmatrix} 1 & -1 \\ 1 & 1 \end{pmatrix}\mathbf{x}$, $\mathbf{x}(0) = \begin{pmatrix} 2 \\ 0 \end{pmatrix}$

\item Find the general solution: $\mathbf{x}' = \begin{pmatrix} -1 & 2 \\ -2 & -1 \end{pmatrix}\mathbf{x}$

\item Solve: $\mathbf{x}' = \begin{pmatrix} 3 & -2 \\ 4 & 3 \end{pmatrix}\mathbf{x}$ with $\mathbf{x}(0) = \begin{pmatrix} 1 \\ 1 \end{pmatrix}$

\item Determine the solution: $\mathbf{x}' = \begin{pmatrix} 0 & 1 \\ -1 & 0 \end{pmatrix}\mathbf{x}$, $\mathbf{x}(0) = \begin{pmatrix} 0 \\ 1 \end{pmatrix}$
\end{enumerate}

\section*{Part D: Stability and Behavior Analysis}

\begin{enumerate}[start=16]
\item Classify the stability of the origin for $\mathbf{x}' = \begin{pmatrix} -2 & 3 \\ -3 & -2 \end{pmatrix}\mathbf{x}$

\item Find the period of oscillation for $\mathbf{x}' = \begin{pmatrix} 0 & -3 \\ 3 & 0 \end{pmatrix}\mathbf{x}$

\item Determine when the solution reaches maximum distance from origin: $\mathbf{x}' = \begin{pmatrix} 1 & -2 \\ 2 & 1 \end{pmatrix}\mathbf{x}$, $\mathbf{x}(0) = \begin{pmatrix} 1 \\ 0 \end{pmatrix}$

\item For what values of $k$ is the origin stable for $\mathbf{x}' = \begin{pmatrix} k & -4 \\ 1 & k \end{pmatrix}\mathbf{x}$?

\item Find a system where solutions spiral inward with period $\pi$.
\end{enumerate}

\section*{Part E: 3\times3 and Higher Dimensional Systems}

\begin{enumerate}[start=21]
\item Solve: $\mathbf{x}' = \begin{pmatrix} 0 & -1 & 0 \\ 1 & 0 & 0 \\ 0 & 0 & 2 \end{pmatrix}\mathbf{x}$

\item Find the general solution: $\mathbf{x}' = \begin{pmatrix} 1 & -1 & 0 \\ 1 & 1 & 0 \\ 0 & 0 & -1 \end{pmatrix}\mathbf{x}$

\item Solve: $\mathbf{x}' = \begin{pmatrix} 0 & 1 & 0 \\ -1 & 0 & 0 \\ 0 & 0 & 0 \end{pmatrix}\mathbf{x}$ with $\mathbf{x}(0) = \begin{pmatrix} 1 \\ 1 \\ 1 \end{pmatrix}$

\item Analyze: $\mathbf{x}' = \begin{pmatrix} -1 & 0 & 2 \\ 0 & -1 & -2 \\ -2 & 2 & -1 \end{pmatrix}\mathbf{x}$

\item Find all eigenvalues: $A = \begin{pmatrix} 0 & -1 & 0 & 0 \\ 1 & 0 & 0 & 0 \\ 0 & 0 & 2 & -3 \\ 0 & 0 & 3 & 2 \end{pmatrix}$
\end{enumerate}

\section*{Part F: Applications and Special Cases}

\begin{enumerate}[start=26]
\item A mass-spring system gives $\mathbf{x}' = \begin{pmatrix} 0 & 1 \\ -4 & 0 \end{pmatrix}\mathbf{x}$ where $x_{1}$ is position, $x_{2}$ is velocity. Find the frequency of oscillation.

\item An RLC circuit yields $\mathbf{x}' = \begin{pmatrix} -1 & -2 \\ 2 & -1 \end{pmatrix}\mathbf{x}$. Determine if the circuit is underdamped, overdamped, or critically damped.

\item Two coupled oscillators give $\mathbf{x}' = \begin{pmatrix} 0 & 1 & 0 & 0 \\ -2 & 0 & 1 & 0 \\ 0 & 0 & 0 & 1 \\ 1 & 0 & -2 & 0 \end{pmatrix}\mathbf{x}$. Find the normal modes.

\item Show that if $A$ is skew-symmetric ($A$^{T}$ = -A$), all eigenvalues are pure imaginary.

\item \textbf{Challenge:} Prove that for the system $\mathbf{x}' = \begin{pmatrix} \alpha & -\beta \\ \beta & \alpha \end{pmatrix}\mathbf{x}$, the solution curves are logarithmic spirals with equation $r = r$_{0e}$^{(\alpha/\beta)\theta}$ in polar coordinates.
\end{enumerate}

\section*{Solutions and Hints}

\textbf{Problem 1:} $\lambda = \pm i$, center (pure rotation)

\textbf{Problem 6:} $\mathbf{x}_{1}(t) = \begin{pmatrix} \cos(2t) \\ \sin(2t) \end{pmatrix}$, $\mathbf{x}_{2}(t) = \begin{pmatrix} \sin(2t) \\ -\cos(2t) \end{pmatrix}$

\textbf{Problem 11:} $\lambda = \pm 2i$, $\mathbf{x}(t) = c_{1}\begin{pmatrix} \cos(2t) \\ \sin(2t)/2 \end{pmatrix} + c_{2}\begin{pmatrix} \sin(2t) \\ -\cos(2t)/2 \end{pmatrix}$

\textbf{Problem 12:} First find $\lambda = 1 \pm i$, then apply initial conditions.

\textbf{Problem 17:} Period $T = 2\pi/\beta = 2\pi/3$

\textbf{Problem 19:} Stable for $k < 0$

\textbf{Problem 26:} Natural frequency $\omega = 2$ rad/s

\textbf{Problem 27:} Underdamped (complex eigenvalues with negative real part)

\textbf{Key Strategy:} Always use Euler's formula to convert complex exponentials to real trig functions. Remember that complex eigenvalues come in conjugate pairs, and you only need to work with one of them.

\textbf{Verification:} Check that your real solutions satisfy the original differential equation!

\end{document}