\documentclass[12pt]{article}
\usepackage{amsmath, amssymb, amsthm, tikz, pgfplots}
\usepackage{geometry, enumitem, mdframed, array, xcolor}
\geometry{margin=1in}

% Custom environments
\newtheorem{definition}{Definition}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{method}{Method}
\newtheorem{example}{Example}
\newmdenv[linecolor=blue,linewidth=2pt]{keypoint}
\newmdenv[linecolor=red,linewidth=2pt]{warning}
\newmdenv[linecolor=green,linewidth=2pt]{insight}
\newmdenv[linecolor=purple,linewidth=2pt]{examtip}
\newmdenv[linecolor=orange,linewidth=2pt]{formula}

\title{ODE Lesson 44: Repeated Roots - Why We Get $t^{k}$ Terms}
\author{ODE 1 with Prof. Adi Ditkowski}
\date{Tel Aviv University}

\begin{document}

\maketitle

\section{Introduction: The Problem of Repeated Roots}

When the characteristic equation of a linear ODE with constant coefficients has repeated roots, the naive approach of using only exponential solutions fails to generate enough linearly independent solutions.

\begin{definition}[Multiplicity of a Root]
If the characteristic polynomial $p(r)$ can be factored as $p(r) = (r - r_{0})^{m} q(r)$ where $q(r_{0}) \neq 0$, then $r_{0}$ is a root of multiplicity $m$.
\end{definition}

\begin{theorem}[Dimension of Solution Space]
An $n$-th order linear homogeneous ODE has an $n$-dimensional solution space. Therefore, we need exactly $n$ linearly independent solutions.
\end{theorem}

\section{Failure of Simple Exponentials}

\begin{example}[The Problem]
Consider $y'' - 4y' + 4y = 0$ with characteristic equation $(r-2)^{2} = 0.

If we only use y = $e^{2t}$, we have just one solution. But we need two linearly independent solutions for this second-order equation.
\end{example}

\begin{keypoint}
When a root r has multiplicity $m > 1, the solution $e^{rt}$ alone cannot span the m-dimensional subspace of solutions corresponding to that root.
\end{keypoint}

\section{Derivation via Reduction of Order}

\begin{method}[Reduction of Order for Repeated Roots]
Given one solution y_{1} = $e^{rt}$ where r is a repeated root, we seek a second solution of the form:
y_{2} = v(t) y_{1} = v(t) $e^{rt}$
\end{method}

\begin{theorem}[Second Solution for Double Root]
If r$ is a double root of the characteristic equation for $y'' + py' + qy = 0, and y_{1} = $e^{rt}$ is one solution, then y_{2} = te^{rt} is a second, linearly independent solution.
\end{theorem}

\begin{proof}
Let y_{2} = v(t)$e^{rt}$. Then:
\begin{align}
y_{2}' &= v'$e^{rt}$ + rve^{rt} \\
y_{2}'' &= v''$e^{rt}$ + 2rv'$e^{rt}$ + r^{2ve}^{rt}
\end{align}

Substituting into the differential equation:
$v''$e^{rt}$ + 2rv'$e^{rt}$ + r^{2ve}^{rt} + p(v'$e^{rt}$ + rve^{rt}) + qve^{rt} = 0

Factoring out $e^{rt}$:
$v'' + v'(2r + p) + v(r^{2} + pr + q) = 0$$

Since $r$ is a double root of $\lambda^{2} + p\lambda + q = 0$:
\begin{enumerate}
\item $r^{2} + pr + q = 0$ (r is a root)
\item $2r + p = 0$ (derivative of characteristic polynomial at r equals 0)
\end{enumerate}

Therefore: $v'' = 0$, giving $v = c_{1t} + c_{2}$.

Taking $c_{1} = 1, c_{2} = 0 yields y_{2} = te^{rt}.
\end{proof}

\section{The Differential Operator Approach}

\begin{insight}
The differential equation (D - r)^{m}[y] = 0 naturally produces solutions t^{k} $e^{rt}$ for k = 0, 1, \ldots, m-1$.
\end{insight}

\begin{theorem}[Operator Factorization]
If the characteristic polynomial factors as $p(\lambda) = (\lambda - r)^{m}$, then the differential operator factors as:
$$L = (D - r)^{m}$$
where $D = \frac{d}{dt}$.
\end{theorem}

\begin{lemma}[Kernel of $(D - r)^{m}$]
The kernel (null space) of the operator $(D - r)^{m}$ is:
$\text{ker}((D - r)^{m}) = \text{span}\{$e^{rt}$, te^{rt}, t^{2e}^{rt}, \ldots, t^{m-1}e^{rt}\}$
\end{lemma}

\begin{proof}[Proof by Induction]
Base case ($m = 1): (D - r)[y] = 0 \Rightarrow y = ce^{rt}.

Inductive step: Assume true for m = k. For $m = k + 1$:
If $(D - r)^{k+1}[y] = 0$, let $w = (D - r)[y]$.
Then $(D - r)^{k}[w] = 0, so w = p_{k}(t)$e^{rt}$ where p_{k} is a polynomial of degree $\leq k-1.

Solving (D - r)[y] = p_{k}(t)$e^{rt}$:
Using integrating factor $e^{-rt}$: \frac{d}{dt}[e^{-rt}y] = p_{k}(t)

Integrating: e^{-rt}y = P_{k+1}(t) + C where $P_{k+1}$ has degree $\leq k$.

Therefore: y = [P_{k+1}(t) + C]$e^{rt}$ = p_{k+1}(t)$e^{rt}$ where \deg(p_{k+1}) \leq k.
\end{proof}

\section{The Limit Perspective}

\begin{theorem}[Repeated Roots as Limits]
The solution te^{rt} can be understood as the limit of solutions for nearby distinct roots.
\end{theorem}

\begin{proof}[Heuristic Derivation]
Consider two roots r_{1} = r and $r_{2} = r + \epsilon$ with solutions:
$y = c_{1} $e^{rt}$ + c_{2} $e^{(r+\epsilon)t}$

Rewriting:
$y = c_{1} $e^{rt}$ + c_{2} $e^{rt}$ e^{\epsilon t}$

For small $\epsilon: $e^{\epsilon t}$ \approx 1 + \epsilon t

Thus:
y \approx (c_{1} + c_{2})$e^{rt}$ + c_{2} \epsilon t $e^{rt}$

As $\epsilon \to 0$, setting $A = c_{1} + c_{2}$ and $B = \lim_{\epsilon \to 0} c_{2} \epsilon$:
$y = Ae^{rt} + Bte^{rt}
\end{proof}

\section{Higher Multiplicities}

\begin{theorem}[General Multiplicity Case]
If $r$ is a root of multiplicity $m$, then the $m$ linearly independent solutions are:
$e^{rt}$, \, te^{rt}, \, t^{2e}^{rt}, \, \ldots, \, t^{m-1}$e^{rt}$
\end{theorem}

\begin{formula}
For a root r$ of multiplicity $m$:
$\text{Solutions} = \{t^{k} $e^{rt}$ : k = 0, 1, 2, \ldots, m-1\}$
\end{formula}

\section{Complex Repeated Roots}

\begin{theorem}[Complex Repeated Roots]
If $\alpha \pm i\beta$ are complex conjugate roots each of multiplicity $m$, the $2m$ real-valued linearly independent solutions are:
$t^{k} $e^{\alpha t}$ \cos(\beta t), \, t^{k} $e^{\alpha t}$ \sin(\beta t) \quad \text{for } k = 0, 1, \ldots, m-1$
\end{theorem}

\begin{warning}
For complex repeated roots, both the sine AND cosine terms need all powers of $t$ up to $t^{m-1}$.
\end{warning}

\section{The Wronskian for Repeated Roots}

\begin{theorem}[Linear Independence via Wronskian]
The functions \{t^{k} $e^{rt}$ : k = 0, 1, \ldots, m-1\} are linearly independent.
\end{theorem}

\begin{proof}[Wronskian Calculation]
For simplicity, consider $m = 2 with solutions $e^{rt}$ and te^{rt}:

W(t) = \begin{vmatrix}
$e^{rt}$ & te^{rt} \\
re^{rt} & (1 + rt)$e^{rt}$
\end{vmatrix}$$

$W(t) = $e^{rt}$ \cdot (1 + rt)$e^{rt}$ - te^{rt} \cdot re^{rt} = $e^{2rt}$(1 + rt - rt) = $e^{2rt}$ \neq 0$

For general $m$, the Wronskian is:
$W(t) = $e^{mrt}$ \cdot \prod_{0 \leq i < j < m} (j - i) \neq 0$
\end{proof}

\section{Connection to Jordan Normal Form}

\begin{insight}
The appearance of $t^{k}$ terms in solutions corresponds exactly to the Jordan block structure of the companion matrix.
\end{insight}

\begin{theorem}[Jordan Form and Solutions]
For a system $\mathbf{x}' = A\mathbf{x}$ where $A$ has a Jordan block:
$$J = \begin{pmatrix}
\lambda & 1 & 0 & \cdots & 0 \\
0 & \lambda & 1 & \cdots & 0 \\
\vdots & \vdots & \ddots & \ddots & \vdots \\
0 & 0 & 0 & \lambda & 1 \\
0 & 0 & 0 & 0 & \lambda
\end{pmatrix}$$

The matrix exponential is:
$e^{Jt}$ = $e^{\lambda t}$ \begin{pmatrix}
1 & t & \frac{t^{2}}{2!} & \cdots & \frac{t^{m-1}}{(m-1)!} \\
0 & 1 & t & \cdots & \frac{t^{m-2}}{(m-2)!} \\
\vdots & \vdots & \ddots & \ddots & \vdots \\
0 & 0 & 0 & 1 & t \\
0 & 0 & 0 & 0 & 1
\end{pmatrix}$$
\end{theorem}

\section{Algorithm for Repeated Roots}

\begin{method}[Complete Solution Construction]
\begin{enumerate}
\item Factor the characteristic polynomial completely.
\item For each distinct root $r_{i}$ with multiplicity $m_{i}$:
    \begin{itemize}
    \item If $r_{i}$ is real: Include solutions t^{k} $e^{r_{i}$ t} for $k = 0, 1, \ldots, m_{i} - 1$
    \item If $r_{i} = \alpha + i\beta$ is complex (with conjugate $\alpha - i\beta$):
        Include solutions t^{k} $e^{\alpha t}$ \cos(\beta t) and t^{k} $e^{\alpha t}$ \sin(\beta t) for $k = 0, 1, \ldots, m_{i} - 1$
    \end{itemize}
\item Form the general solution as a linear combination of all these functions.
\end{enumerate}
\end{method}

\section{Worked Examples}

\begin{example}[Triple Root]
Solve: $y''' - 3y'' + 3y' - y = 0$

\textbf{Solution:}
Characteristic equation: $r^{3} - 3r^{2} + 3r - 1 = 0$

Recognize this as $(r-1)^{3} = 0$ (expand to verify).

Root: $r = 1 with multiplicity 3.

Solutions: $e^{t}$, te^{t}, t^{2e}^t

General solution: y(t) = (c_{1} + c_{2} t + c_{3} t^{2})$e^{t}$
\end{example}

\begin{example}[Complex Double Roots]
Solve: y^{(4)} + 8y'' + 16y = 0$

\textbf{Solution:}
Characteristic equation: $r^{4} + 8r^{2} + 16 = 0$

This is $(r^{2} + 4)^{2} = 0$, giving $(r - 2i)^{2}(r + 2i)^{2} = 0$

Roots: $\pm 2i$, each with multiplicity 2.

Real solutions:
\begin{itemize}
\item From $2i$ (mult. 2): $\cos(2t), t\cos(2t), \sin(2t), t\sin(2t)$
\end{itemize}

General solution: $y(t) = (c_{1} + c_{2} t)\cos(2t) + (c_{3} + c_{4} t)\sin(2t)$
\end{example}

\begin{examtip}
Prof. Ditkowski often presents the characteristic polynomial already factored. Don't waste time trying to factor what's already factored - just identify multiplicities and write solutions systematically.
\end{examtip}

\end{document}