\documentclass[12pt]{article}
\usepackage{amsmath, amssymb, enumitem}
\usepackage{geometry}
\geometry{margin=1in}

\title{Practice Problems: Direct Integration}
\author{Lesson 11 Exercises}

\begin{document}
\maketitle

\section*{Part A: Recognition and Classification (6 problems)}

\begin{enumerate}
\item Which of the following ODEs can be solved by direct integration?
    \begin{enumerate}
        \item $y' = x^2 + y$
        \item $y' = e^x \cos(x)$
        \item $y' = \frac{y}{x}$
        \item $y'' = \sin(x) - 2$
    \end{enumerate}

\item For the ODE $y^{(4)} = 0$, what is the general form of the solution?

\item How many arbitrary constants appear in the general solution of $\frac{d^5y}{dx^5} = e^x$?

\item True or False: If $y'' = f(x)$ has solution $y = g(x)$, then $y = g(x) + Ax + B$ is also a solution for any constants $A, B$.

\item What is the minimum number of initial conditions needed to uniquely determine the solution of $y''' = x^2$?

\item If $dy/dx = f(x)$ and $f(x)$ is odd, what can you say about the solution $y(x)$ passing through the origin?
\end{enumerate}

\section*{Part B: Basic Integration Problems (6 problems)}

\begin{enumerate}[resume]
\item Solve: $\frac{dy}{dx} = 3x^2 - 4x + 1$

\item Solve: $\frac{dy}{dx} = \frac{1}{x} + \frac{1}{x^2}$ for $x > 0$

\item Find the general solution: $y' = e^{2x} + \sin(3x)$

\item Solve: $\frac{dy}{dx} = \frac{1}{\sqrt{1-x^2}}$ for $|x| < 1$

\item Find $y(x)$ if $y' = \sec^2(x)$ and $y(0) = 2$

\item Solve: $\frac{dy}{dx} = |x|$ (consider both $x \geq 0$ and $x < 0$)
\end{enumerate}

\section*{Part C: Higher-Order Direct Integration (6 problems)}

\begin{enumerate}[resume]
\item Solve: $y'' = 6x$ with $y(0) = 1$ and $y'(0) = -1$

\item Find the general solution: $\frac{d^3y}{dx^3} = 24$

\item Solve: $y'' = \cos(x)$ with $y(0) = 0$ and $y(\pi) = 0$

\item Find $y(x)$ if $y''' = e^x$ with $y(0) = y'(0) = y''(0) = 1$

\item Solve: $\frac{d^4y}{dx^4} = 0$ with $y(0) = 1$, $y'(0) = 0$, $y''(0) = 2$, $y'''(0) = 0$

\item A particle moves with constant jerk (rate of change of acceleration) $j = 2$ m/s³. If at $t = 0$ the particle is at rest at the origin with zero acceleration, find its position at time $t$.
\end{enumerate}

\section*{Part D: Initial Value Problems with Definite Integrals (5 problems)}

\begin{enumerate}[resume]
\item Using the definite integral method, solve: $y' = x^2$ with $y(1) = 3$

\item Express the solution to $y' = \sin(x^2)$ with $y(0) = 1$ as a definite integral.

\item Solve: $y' = \frac{1}{1+x^2}$ with $y(0) = \pi/4$

\item Find $y(2)$ if $y' = e^{-x^2}$ and $y(0) = 0$ (express as an integral if needed).

\item Show that the solution to $y' = f(x)$ with $y(a) = A$ and $y(b) = B$ must satisfy: $B - A = \int_a^b f(x)\,dx$
\end{enumerate}

\section*{Part E: Mixed Exam-Style Questions (7 problems)}

\begin{enumerate}[resume]
\item \textbf{[Prof. Ditkowski Special]} Consider $y'' = x\sin(x)$.
    \begin{enumerate}
        \item Find the general solution
        \item Find the particular solution with $y(0) = 1$ and $y'(\pi) = 0$
        \item Verify your solution
    \end{enumerate}

\item \textbf{[Conceptual]} If $y_1$ is a solution to $y'' = f(x)$ and $y_2$ is a solution to $y'' = g(x)$, what ODE does $y_1 + y_2$ solve?

\item \textbf{[Multi-part]} A projectile is launched vertically with $y'' = -10$ m/s² (taking up as positive).
    \begin{enumerate}
        \item Find the general solution for height $y(t)$
        \item If launched from ground level with initial velocity 30 m/s, find $y(t)$
        \item When does it reach maximum height?
        \item What is the maximum height?
    \end{enumerate}

\item \textbf{[Verification Focus]} A student claims that $y = x^3/3 + \ln|x| + 2x + 5$ solves $y' = x^2 + 1/x + 2$ with $y(1) = 5 + 7/3$. Verify or disprove this claim.

\item \textbf{[Domain Awareness]} Solve $y' = \frac{1}{x\ln(x)}$ for $x > 1$ with $y(e) = 0$. State the domain of your solution.

\item \textbf{[Integration Challenge]} Find all functions $y(x)$ such that $y''' = y$ and $y(0) = y'(0) = y''(0) = 0$.

\item \textbf{[Prof. Ditkowski Comprehensive]}
    Consider the ODE: $\frac{d^2y}{dx^2} = \frac{2}{(1+x^2)^2}$
    \begin{enumerate}
        \item Show that $y' = \frac{x}{1+x^2} + C_1$ is the first integral
        \item Find the general solution
        \item Find the solution satisfying $y(0) = 0$ and $\lim_{x \to \infty} y'(x) = 1$
        \item Sketch the solution curve
    \end{enumerate}
\end{enumerate}

\section*{Answer Key with Essential Hints}

\textbf{1.} Only (b) and (d) can be solved by direct integration

\textbf{2.} $y = C_1x^3 + C_2x^2 + C_3x + C_4$ (cubic polynomial)

\textbf{5.} Three initial conditions needed

\textbf{7.} $y = x^3 - 2x^2 + x + C$

\textbf{9.} $y = \frac{1}{2}e^{2x} - \frac{1}{3}\cos(3x) + C$

\textbf{13.} $y = x^3 - x + 1$

\textbf{14.} $y = 4x^3 + C_1x^2 + C_2x + C_3$

\textbf{18.} $y = \frac{t^3}{3}$ (position function)

\textbf{19.} $y = \frac{x^3}{3} - \frac{1}{3} + 3 = \frac{x^3}{3} + \frac{8}{3}$

\textbf{22.} $y(2) = \int_0^2 e^{-t^2}dt$ (cannot be expressed in elementary functions)

\textbf{24.} Use integration by parts: $\int x\sin(x)dx = -x\cos(x) + \sin(x) + C$

\textbf{26.} Max height at $t = 3$ seconds, height = 45 meters

\textbf{28.} Solution: $y = \ln|\ln(x)|$, domain: $x > 1$, $x \neq e$

\textbf{30.} First show that $\int \frac{2dx}{(1+x^2)^2} = \frac{x}{1+x^2} + C$ using substitution

\end{document}