\documentclass[12pt]{article}
\usepackage{amsmath, amssymb, amsthm, enumitem, geometry}
\geometry{margin=1in}

\title{Lesson 30: Practice Problems - Distinct Eigenvalues}
\author{ODE 1 - Prof. Adi Ditkowski}
\date{}

\usepackage{mdframed}
\usepackage{xcolor}
\usepackage{tikz}
\begin{document}
\maketitle

\section*{Part A: Eigenvalue/Eigenvector Computation}

\begin{enumerate}
\item Find all eigenvalues and eigenvectors of $A = \begin{pmatrix} 3 & 1 \\ 2 & 4 \end{pmatrix}$

\item Find all eigenvalues and eigenvectors of $A = \begin{pmatrix} 5 & -2 \\ 3 & -2 \end{pmatrix}$

\item Find all eigenvalues and eigenvectors of $A = \begin{pmatrix} 1 & 2 & 0 \\ 0 & 3 & 0 \\ 0 & 0 & -1 \end{pmatrix}$

\item Verify that $\mathbf{v} = \begin{pmatrix} 1 \\ 2 \end{pmatrix}$ is an eigenvector of $A = \begin{pmatrix} 4 & 1 \\ 2 & 5 \end{pmatrix}$ and find the corresponding eigenvalue.

\item For the matrix $A = \begin{pmatrix} a & 1 \\ 0 & b \end{pmatrix}$, find conditions on $a$ and $b$ for distinct real eigenvalues.
\end{enumerate}

\section*{Part B: 2\times2 Systems with Distinct Real Eigenvalues}

\begin{enumerate}[start=6]
\item Solve the system $\mathbf{x}' = \begin{pmatrix} 2 & 1 \\ 3 & 4 \end{pmatrix}\mathbf{x}$

\item Solve the IVP: $\mathbf{x}' = \begin{pmatrix} 1 & 2 \\ 2 & 1 \end{pmatrix}\mathbf{x}$, $\mathbf{x}(0) = \begin{pmatrix} 3 \\ 1 \end{pmatrix}$

\item Find the solution to $\mathbf{x}' = \begin{pmatrix} 4 & -2 \\ 1 & 1 \end{pmatrix}\mathbf{x}$ that satisfies $x_{1}(0) = 2, x_{2}(0) = -1$

\item Determine all solutions of $\mathbf{x}' = \begin{pmatrix} 3 & 4 \\ 1 & 3 \end{pmatrix}\mathbf{x}$ that remain bounded as $t \to \infty$

\item Find the fundamental matrix for $\mathbf{x}' = \begin{pmatrix} 5 & 3 \\ 1 & 3 \end{pmatrix}\mathbf{x}$
\end{enumerate}

\section*{Part C: 3\times3 Systems with Distinct Eigenvalues}

\begin{enumerate}[start=11]
\item Solve: $\mathbf{x}' = \begin{pmatrix} 1 & 0 & 0 \\ 0 & 2 & 0 \\ 0 & 0 & 3 \end{pmatrix}\mathbf{x}$

\item Solve: $\mathbf{x}' = \begin{pmatrix} 2 & 1 & 1 \\ 0 & 3 & 1 \\ 0 & 0 & 4 \end{pmatrix}\mathbf{x}$

\item Find the general solution: $\mathbf{x}' = \begin{pmatrix} 0 & 1 & 0 \\ 0 & 0 & 1 \\ 6 & -11 & 6 \end{pmatrix}\mathbf{x}$

\item Solve the IVP: $\mathbf{x}' = \begin{pmatrix} 1 & 0 & 1 \\ 0 & 2 & 0 \\ 1 & 0 & 1 \end{pmatrix}\mathbf{x}$, $\mathbf{x}(0) = \begin{pmatrix} 1 \\ 1 \\ 1 \end{pmatrix}$

\item Find all equilibrium solutions and their stability for: $\mathbf{x}' = \begin{pmatrix} -1 & 0 & 0 \\ 0 & -2 & 0 \\ 0 & 0 & 3 \end{pmatrix}\mathbf{x}$
\end{enumerate}

\section*{Part D: Stability and Behavior Analysis}

\begin{enumerate}[start=16]
\item For what values of $k$ is the origin a stable equilibrium for $\mathbf{x}' = \begin{pmatrix} -1 & k \\ 0 & -2 \end{pmatrix}\mathbf{x}$?

\item Classify the equilibrium at the origin for $\mathbf{x}' = \begin{pmatrix} 3 & 5 \\ 1 & -1 \end{pmatrix}\mathbf{x}$

\item Find a system $\mathbf{x}' = A\mathbf{x}$ where all solutions approach the line $x_{1} = x_{2}$ as $t \to \infty$

\item Determine the long-term behavior of solutions to $\mathbf{x}' = \begin{pmatrix} -3 & 2 \\ 1 & -2 \end{pmatrix}\mathbf{x}$

\item Show that if $A$ has all negative eigenvalues, then $\|\mathbf{x}(t)\| \to 0$ as $t \to \infty$
\end{enumerate}

\section*{Part E: Special Cases and Theory}

\begin{enumerate}[start=21]
\item Prove that if $A$ is symmetric, all eigenvalues are real

\item Show that $\text{tr}(A) = \sum \lambda$_{i}$$ and $\det(A) = \prod \lambda$_{i}$$

\item If $A$ has eigenvalues 2, 3, 5, what are the eigenvalues of $A^{2}$? Of $A^{-1}$?

\item Construct a 2\times2 matrix with eigenvalues $\lambda_{1} = 1, \lambda_{2} = -2$ and eigenvectors $\mathbf{v}_{1} = \begin{pmatrix} 1 \\ 1 \end{pmatrix}, \mathbf{v}_{2} = \begin{pmatrix} 1 \\ -1 \end{pmatrix}$

\item Find a 3\times3 upper triangular matrix with eigenvalues 1, 2, 3 and determine its eigenvectors
\end{enumerate}

\section*{Part F: Application Problems}

\begin{enumerate}[start=26]
\item Two tanks contain salt solutions. Tank 1 has rate of change $x_{1}' = -0.1x_{1} + 0.05x_{2}$ and Tank 2 has $x_{2}' = 0.1x_{1} - 0.15x_{2}$. Find the salt amounts over time if initially $x_{1}(0) = 100, x_{2}(0) = 50$.

\item A predator-prey model gives $\mathbf{x}' = \begin{pmatrix} 2 & -1 \\ 1 & -1 \end{pmatrix}\mathbf{x}$ where $x_{1}$ is prey, $x_{2}$ is predator population. Analyze the long-term behavior.

\item Coupled springs lead to $\mathbf{x}' = \begin{pmatrix} 0 & 1 \\ -2 & 0 \end{pmatrix}\mathbf{x} + \begin{pmatrix} 0 & 0 \\ 0 & -3 \end{pmatrix}\mathbf{x}$. Find the general solution.

\item An electrical circuit gives $\mathbf{x}' = \begin{pmatrix} -R/L & -1/L \\ 1/C & 0 \end{pmatrix}\mathbf{x}$ with $R = 2, L = 1, C = 0.5$. Solve for the current and voltage.

\item \textbf{Challenge:} Show that the solution to $\mathbf{x}' = A\mathbf{x}$ can be written as $\mathbf{x}(t) = $e^{At}\mathbf{x}_{0$}$ and verify this for a diagonal matrix.
\end{enumerate}

\section*{Solutions and Hints}

\textbf{Problem 1:} $\lambda_{1} = 5, \lambda_{2} = 2$; $\mathbf{v}_{1} = \begin{pmatrix} 1 \\ 2 \end{pmatrix}, \mathbf{v}_{2} = \begin{pmatrix} 1 \\ -1 \end{pmatrix}$

\textbf{Problem 7:} First find $\lambda_{1} = 3, \lambda_{2} = -1$. General solution involves $e^{3t}$ and $e^{-t}$ terms.

\textbf{Problem 11:} Diagonal matrix - eigenvectors are standard basis vectors.

\textbf{Problem 16:} Stable for all $k$ since both eigenvalues are negative regardless of $k$.

\textbf{Problem 22:} Use the fact that $\text{tr}(A) = \text{tr}(PDP^{-1}) = \text{tr}(D)$ where $D$ is diagonal with eigenvalues.

\textbf{Key Strategy:} Always verify your eigenvalues by checking that $\det(A - \lambda I) = 0$ and eigenvectors by confirming $A\mathbf{v} = \lambda\mathbf{v}$.

\end{document}