\documentclass[12pt]{article}
\usepackage{amsmath, amssymb, enumitem}
\usepackage{geometry}
\geometry{margin=1in}

\title{Practice Problems: Lesson 10 - Qualitative Analysis}
\author{Master qualitative techniques without solving!}
\date{}

\begin{document}
\maketitle

\section*{Part A: Basic Concepts (6 problems)}

\begin{enumerate}
\item Draw the phase line for $\frac{dy}{dx} = y^2 - 4$ and classify all equilibria.

\item Using the derivative test, determine stability of equilibria for $\frac{dy}{dx} = \sin(y)$.

\item For $\frac{dy}{dx} = e^y - 1$, find the equilibrium and determine its stability without computing derivatives.

\item Explain why solutions to $\frac{dy}{dx} = y^2 + 1$ cannot have equilibria.

\item If $y_1(0) = 1$ and $y_2(0) = 2$ are solutions to $\frac{dy}{dx} = -y$, what is their relationship for all $x > 0$?

\item For $\frac{dy}{dx} = f(y)$ where $f$ is strictly decreasing, prove all equilibria are stable.
\end{enumerate}

\section*{Part B: Core Techniques (6 problems)}

\begin{enumerate}[resume]
\item Analyze the complete phase portrait for $\frac{dy}{dx} = y^3 - y$.

\item For $\frac{dy}{dx} = (y-1)(y-2)(y-3)$, determine the long-term behavior of solutions with different initial conditions.

\item Find regions where solutions to $\frac{dy}{dx} = x^2 - y$ are concave up vs concave down.

\item Using comparison, prove that solutions to $\frac{dy}{dx} = -y + \sin(x)$ are bounded.

\item For $\frac{dy}{dx} = y - y^3$, find the inflection points of solution curves.

\item Determine all asymptotic behaviors for $\frac{dy}{dx} = y^2(1-y)$.
\end{enumerate}

\section*{Part C: Applications (5 problems)}

\begin{enumerate}[resume]
\item A population model follows $\frac{dP}{dt} = P(2 - P)(P - 1)$. Analyze long-term behavior for all initial populations.

\item For the Gompertz equation $\frac{dy}{dx} = y\ln(K/y)$ with $K > 0$, analyze stability without solving.

\item Show that $\frac{dy}{dx} = x + y^2$ has no periodic solutions.

\item Find a first integral for $\frac{dy}{dx} = \frac{2xy}{x^2 - y^2}$.

\item For $\frac{dy}{dx} = -y + e^{-x}$, determine $\lim_{x \to \infty} y(x)$ without solving.
\end{enumerate}

\section*{Part D: Advanced/Theoretical (5 problems)}

\begin{enumerate}[resume]
\item Prove that if $H(x,y)$ is a first integral, then $\nabla H \perp (1, f(x,y))$ at each point.

\item Show that for $\frac{dy}{dx} = f(y)g(x)$ with $g(x) > 0$, stability is determined solely by $f$.

\item For what functions $f(y)$ does $\frac{dy}{dx} = f(y)$ have exactly three equilibria with alternating stability?

\item Prove that if all solutions to $\frac{dy}{dx} = f(x,y)$ are bounded, then $f$ must change sign.

\item Characterize all ODEs of the form $\frac{dy}{dx} = f(x)g(y)$ that admit periodic solutions.
\end{enumerate}

\section*{Part E: Exam-Style Questions (6 problems)}

\begin{enumerate}[resume]
\item \textbf{[Prof. Ditkowski Special]} For $\frac{dy}{dx} = y^2 - xy$:
    \begin{enumerate}[label=\alph*)]
        \item Find all curves where solutions have horizontal tangents
        \item Determine regions of increasing/decreasing behavior
        \item Analyze concavity
        \item Describe all possible asymptotic behaviors
    \end{enumerate}

\item Given $\frac{dy}{dx} = (1-y^2)(x^2 + 1)$:
    \begin{enumerate}[label=\alph*)]
        \item Find and classify all equilibria
        \item Prove all solutions are bounded
        \item Determine if periodic solutions exist
        \item Sketch the qualitative behavior
    \end{enumerate}

\item \textbf{[Multi-method]} For $\frac{dy}{dx} = y(1-y)(2-y)$:
    \begin{enumerate}[label=\alph*)]
        \item Draw the complete phase line
        \item Use linearization to confirm stability
        \item Find maximum growth rate locations
        \item Describe basin of attraction for each stable equilibrium
    \end{enumerate}

\item Without solving, prove that all solutions to $\frac{dy}{dx} = -x - y^3$ approach zero as $x \to \infty$.

\item \textbf{[Energy Method]} Show that $\frac{dy}{dx} = \frac{x^3 - xy^2}{x^2y + y^3}$ has circular solution curves by finding an appropriate first integral.

\item A system satisfies $\frac{dy}{dx} = f(y)$ where $f$ is continuous, $f(0) = f(1) = f(2) = 0$, $f'(0) > 0$, $f'(1) < 0$, $f'(2) > 0$.
    \begin{enumerate}[label=\alph*)]
        \item Sketch the phase line
        \item How many distinct asymptotic behaviors exist?
        \item What fraction of initial conditions lead to $y \to 1$?
        \item Can solutions oscillate?
    \end{enumerate}
\end{enumerate}

\section*{Answer Key with Hints}

\textbf{Problem 1:} Equilibria at $y = \pm 2$; $y = -2$ unstable, $y = 2$ stable

\textbf{Problem 7:} Three equilibria: $y = 0$ (unstable), $y = \pm 1$ (stable)

\textbf{Problem 10:} Use comparison with $\frac{dy}{dx} = -y \pm 1$

\textbf{Problem 16:} First integral: $H = x^2y - y^3/3$

\textbf{Problem 25:} Use Lyapunov function $V = x^2 + y^2$

\textbf{Problem 28c:} Basin of $y=1$ includes all $y \in (0,2)$

\end{document}