\documentclass[12pt]{article}
\usepackage{amsmath, amssymb, amsthm, enumitem, mdframed}
\usepackage{geometry, xcolor}
\geometry{margin=1in}

\newmdenv[linecolor=blue,linewidth=2pt]{hint}
\newtheorem{problem}{Problem}

\title{Lesson 16: Practice Problems - Variation of Constants}
\author{ODE 1 - Prof. Adi Ditkowski}
\date{Tel Aviv University}

\begin{document}
\maketitle

\section*{Part A: Basic Variation of Constants (5 Problems)}

\begin{problem}
Solve $y' - 2y = e^{4t}$ using variation of constants.
\end{problem}

\begin{problem}
Solve $y' + y = \sin(2t)$ using variation of constants.
\end{problem}

\begin{problem}
Solve $y' + 3y = t^2$ using variation of constants.
\end{problem}

\begin{problem}
Solve $y' - \frac{1}{t}y = t^3$ for $t > 0$ using variation of constants.
\end{problem}

\begin{problem}
Solve $y' + 4y = e^{-4t}$ using variation of constants. Note the resonance!
\end{problem}

\section*{Part B: Non-Standard Forcing Functions (6 Problems)}

\begin{problem}
Solve $y' - y = e^t \ln(t)$ for $t > 0$.
\end{problem}

\begin{problem}
Solve $y' + 2y = \frac{\sin(t)}{t}$ for $t > 0$.
\end{problem}

\begin{problem}
Solve $y' + \frac{1}{t}y = \cos(\ln(t))$ for $t > 0$.
\end{problem}

\begin{problem}
Solve $y' - 3y = e^{3t}t^2$ (resonance with polynomial).
\end{problem}

\begin{problem}
Solve $y' + y = \frac{1}{1+e^t}$.
\end{problem}

\begin{problem}
Solve $y' + ty = te^{-t^2/2}$.
\end{problem}

\section*{Part C: Initial Value Problems (5 Problems)}

\begin{problem}
Solve $y' - 2y = e^t$, $y(0) = 3$.
\end{problem}

\begin{problem}
Solve $y' + \frac{2}{t}y = t^2$, $y(1) = 2$ for $t > 0$.
\end{problem}

\begin{problem}
Solve $y' + 3y = e^{-3t}\cos(t)$, $y(0) = 0$.
\end{problem}

\begin{problem}
Solve $y' - y = \frac{1}{1+e^{-t}}$, $y(0) = 1$.
\end{problem}

\begin{problem}
Solve $y' + (\tan t)y = \sec t$, $y(0) = 1$ for $-\pi/2 < t < \pi/2$.
\end{problem}

\section*{Part D: Comparison with Other Methods (5 Problems)}

\begin{problem}
Solve $y' + 2y = 3e^{-2t}$ using:
a) Integrating factor method
b) Variation of constants
c) Show the solutions are identical
\end{problem}

\begin{problem}
For $y' - y = e^t$:
a) Explain why undetermined coefficients fails
b) Solve using variation of constants
c) Interpret the resonance physically
\end{problem}

\begin{problem}
Solve $y' + p(t)y = g(t)$ where:
\[p(t) = \begin{cases} 1 & 0 \leq t < 1 \\ 2 & t \geq 1 \end{cases}, \quad g(t) = 1\]
Use variation of constants on each interval.
\end{problem}

\begin{problem}
Show that for $y' + p(t)y = g(t)$, the variation of constants formula and integrating factor formula give the same result.
\end{problem}

\begin{problem}
Solve $ty' - y = t^2e^t$ for $t > 0$:
a) First convert to standard form
b) Apply variation of constants
c) Verify the solution
\end{problem}

\section*{Part E: Theoretical Problems (4 Problems)}

\begin{problem}
Prove that if $y_1$ and $y_2$ are solutions to $y' + p(t)y = g_1(t)$ and $y' + p(t)y = g_2(t)$ respectively, then $y_1 + y_2$ solves $y' + p(t)y = g_1(t) + g_2(t)$.
\end{problem}

\begin{problem}
Show that the variation of constants formula:
\[y_p = y_h(t) \int \frac{g(t)}{y_h(t)}dt\]
can be written as:
\[y_p(t) = \int_{t_0}^t \frac{y_h(t)}{y_h(s)} g(s) ds\]
Interpret this as a Green's function representation.
\end{problem}

\begin{problem}
Prove that for the equation $y' + p(t)y = g(t)$, if $p(t)$ and $g(t)$ are continuous on $[a,b]$, then the solution exists and is unique on the entire interval.
\end{problem}

\begin{problem}
Derive the variation of constants formula starting from the integrating factor approach. Show all steps.
\end{problem}

\section*{Part F: Exam-Style Problems (5 Problems)}

\begin{problem}[15 points]
Consider the equation $y' + \frac{2t}{1+t^2}y = \frac{1}{1+t^2}$.
\begin{enumerate}
    \item Find the homogeneous solution
    \item Use variation of constants to find the general solution
    \item Find the solution satisfying $y(0) = 1$
    \item Determine the behavior as $t \to \infty$
\end{enumerate}
\end{problem}

\begin{problem}[20 points]
The current $i(t)$ in an RL circuit satisfies:
\[L\frac{di}{dt} + Ri = V(t)\]
where $L = 1$ H, $R = 2$ $\Omega$, and $V(t) = 10e^{-2t}\sin(3t)$ V.
\begin{enumerate}
    \item Convert to standard form
    \item Find the steady-state current using variation of constants
    \item If $i(0) = 0$, find the complete solution
    \item Identify transient and steady-state components
\end{enumerate}
\end{problem}

\begin{problem}[15 points]
For the equation $y' - \frac{n}{t}y = t^m$ where $n, m$ are constants and $t > 0$:
\begin{enumerate}
    \item Find the general solution using variation of constants
    \item For what values of $n$ and $m$ does the solution remain bounded as $t \to \infty$?
    \item Find the particular solution that remains finite at $t = 0$ when possible
\end{enumerate}
\end{problem}

\begin{problem}[20 points]
Consider $y' + y = f(t)$ where:
\[f(t) = \begin{cases}
1 & 0 \leq t < \pi \\
\sin(t) & t \geq \pi
\end{cases}\]
\begin{enumerate}
    \item Solve on $[0, \pi)$ with $y(0) = 0$
    \item Use continuity at $t = \pi$ to find initial condition for $t \geq \pi$
    \item Solve on $[\pi, \infty)$
    \item Sketch the solution
\end{enumerate}
\end{problem}

\begin{problem}[15 points - Theoretical]
\begin{enumerate}
    \item State the variation of constants formula for $y' + p(t)y = g(t)$
    \item Prove that this formula gives a particular solution
    \item Show that if $g(t) = g_1(t) + g_2(t)$, then $y_p = y_{p1} + y_{p2}$
    \item Explain why variation of constants works for any continuous $g(t)$
\end{enumerate}
\end{problem}

\begin{hint}
\textbf{Solution Hints:}
\begin{itemize}
    \item Problem 5: Resonance occurs when forcing matches natural frequency
    \item Problem 10: Use substitution $u = 1 + e^t$
    \item Problem 11: The coefficient $t$ suggests $y_h = Ce^{-t^2/2}$
    \item Problem 20: Convert using $p(t) = 2t/(1+t^2)$, notice $\int p(t)dt = \ln(1+t^2)$
    \item Problem 24: Use piecewise approach with continuity condition
\end{itemize}
\end{hint}

\end{document}