\documentclass[12pt]{article}
\usepackage{amsmath, amssymb, amsthm, enumitem, geometry}
\geometry{margin=1in}

\title{Lesson 31: Practice Problems - Repeated Eigenvalues and Jordan Forms}
\author{ODE 1 - Prof. Adi Ditkowski}
\date{}

\usepackage{mdframed}
\usepackage{xcolor}
\usepackage{tikz}
\begin{document}
\maketitle

\section*{Part A: Identifying Defective Eigenvalues}

\begin{enumerate}
\item Determine the algebraic and geometric multiplicities of each eigenvalue for:
$A = \begin{pmatrix} 2 & 1 \\ 0 & 2 \end{pmatrix}$

\item Find the defect of each eigenvalue for:
$A = \begin{pmatrix} 3 & 1 & 0 \\ 0 & 3 & 0 \\ 0 & 0 & 3 \end{pmatrix}$

\item Show that $A = \begin{pmatrix} 1 & 2 \\ 0 & 1 \end{pmatrix}$ is defective and find its Jordan form.

\item Determine which of these matrices are diagonalizable:
a) $\begin{pmatrix} 2 & 1 \\ 0 & 2 \end{pmatrix}$
b) $\begin{pmatrix} 2 & 0 \\ 0 & 2 \end{pmatrix}$
c) $\begin{pmatrix} 2 & 1 \\ 1 & 2 \end{pmatrix}$

\item For $A = \begin{pmatrix} 4 & 1 & 0 \\ 0 & 4 & 1 \\ 0 & 0 & 4 \end{pmatrix}$, find the size of the largest Jordan block.
\end{enumerate}

\section*{Part B: Finding Generalized Eigenvectors}

\begin{enumerate}[start=6]
\item Find all eigenvectors and generalized eigenvectors for:
$A = \begin{pmatrix} 3 & 1 \\ 0 & 3 \end{pmatrix}$

\item Construct a complete Jordan chain for:
$A = \begin{pmatrix} 2 & 1 & 0 \\ 0 & 2 & 1 \\ 0 & 0 & 2 \end{pmatrix}$

\item Find the Jordan basis for:
$A = \begin{pmatrix} 5 & 4 \\ -1 & 1 \end{pmatrix}$

\item Determine all generalized eigenvectors for:
$A = \begin{pmatrix} 1 & 1 & 0 \\ 0 & 1 & 0 \\ 0 & 0 & 2 \end{pmatrix}$

\item Find a matrix $P$ such that $P^{-1}AP$ is in Jordan form for:
$A = \begin{pmatrix} 4 & 1 \\ -4 & 0 \end{pmatrix}$
\end{enumerate}

\section*{Part C: 2\times2 Systems with Repeated Eigenvalues}

\begin{enumerate}[start=11]
\item Solve: $\mathbf{x}' = \begin{pmatrix} 2 & 1 \\ 0 & 2 \end{pmatrix}\mathbf{x}$

\item Solve the IVP: $\mathbf{x}' = \begin{pmatrix} 3 & 1 \\ -1 & 1 \end{pmatrix}\mathbf{x}$, $\mathbf{x}(0) = \begin{pmatrix} 1 \\ 0 \end{pmatrix}$

\item Find the general solution: $\mathbf{x}' = \begin{pmatrix} -1 & 1 \\ 0 & -1 \end{pmatrix}\mathbf{x}$

\item Solve: $\mathbf{x}' = \begin{pmatrix} 4 & 2 \\ -2 & 0 \end{pmatrix}\mathbf{x}$ with $\mathbf{x}(0) = \begin{pmatrix} 3 \\ -1 \end{pmatrix}$

\item Find all solutions that remain bounded as $t \to \infty$ for:
$\mathbf{x}' = \begin{pmatrix} -2 & 1 \\ 0 & -2 \end{pmatrix}\mathbf{x}$
\end{enumerate}

\section*{Part D: 3\times3 Systems with Jordan Blocks}

\begin{enumerate}[start=16]
\item Solve: $\mathbf{x}' = \begin{pmatrix} 1 & 1 & 0 \\ 0 & 1 & 1 \\ 0 & 0 & 1 \end{pmatrix}\mathbf{x}$

\item Find the general solution: $\mathbf{x}' = \begin{pmatrix} 2 & 1 & 0 \\ 0 & 2 & 0 \\ 0 & 0 & 3 \end{pmatrix}\mathbf{x}$

\item Solve the IVP: $\mathbf{x}' = \begin{pmatrix} 0 & 1 & 0 \\ 0 & 0 & 1 \\ 0 & 0 & 0 \end{pmatrix}\mathbf{x}$, $\mathbf{x}(0) = \begin{pmatrix} 1 \\ 1 \\ 1 \end{pmatrix}$

\item Find the solution: $\mathbf{x}' = \begin{pmatrix} 3 & 1 & 0 \\ 0 & 3 & 1 \\ 0 & 0 & 3 \end{pmatrix}\mathbf{x}$ with $x_{1}(0) = 1, x_{2}(0) = 0, x_{3}(0) = 0$

\item Determine the Jordan form and solve: $\mathbf{x}' = \begin{pmatrix} 2 & 0 & 1 \\ 0 & 2 & 0 \\ 0 & 0 & 2 \end{pmatrix}\mathbf{x}$
\end{enumerate}

\section*{Part E: Mixed Eigenvalue Problems}

\begin{enumerate}[start=21]
\item Solve: $\mathbf{x}' = \begin{pmatrix} 1 & 1 & 0 \\ 0 & 1 & 0 \\ 0 & 0 & 2 \end{pmatrix}\mathbf{x}$ (repeated and distinct eigenvalues)

\item Find the general solution: $\mathbf{x}' = \begin{pmatrix} 3 & 1 & 0 & 0 \\ 0 & 3 & 0 & 0 \\ 0 & 0 & -1 & 1 \\ 0 & 0 & 0 & -1 \end{pmatrix}\mathbf{x}$

\item Solve: $\mathbf{x}' = \begin{pmatrix} 2 & 1 & 0 \\ 0 & 2 & 0 \\ 0 & 0 & -3 \end{pmatrix}\mathbf{x}$ with $\mathbf{x}(0) = \begin{pmatrix} 1 \\ 1 \\ 1 \end{pmatrix}$

\item Analyze stability for: $\mathbf{x}' = \begin{pmatrix} -2 & 1 & 0 \\ 0 & -2 & 1 \\ 0 & 0 & -2 \end{pmatrix}\mathbf{x}$

\item Find the fundamental matrix: $\mathbf{x}' = \begin{pmatrix} 0 & 1 & 0 \\ 0 & 0 & 0 \\ 0 & 0 & 1 \end{pmatrix}\mathbf{x}$
\end{enumerate}

\section*{Part F: Theory and Applications}

\begin{enumerate}[start=26]
\item Prove that if $\lambda$ is a repeated eigenvalue with full defect $n-1$, then the solutions contain terms up to $t^{n-1}$e^{\lambda$ t}$.

\item Show that for a 2\times2 matrix with repeated eigenvalue $\lambda$, the trace equals $2\lambda$ and the determinant equals $\lambda^{2}$.

\item A coupled system has matrix $A = \begin{pmatrix} -1 & 1 \\ 0 & -1 \end{pmatrix}$. Find the time when the solution with $\mathbf{x}(0) = \begin{pmatrix} 0 \\ 1 \end{pmatrix}$ reaches its maximum norm.

\item For the radioactive decay chain with matrix $A = \begin{pmatrix} -k & 0 & 0 \\ k & -k & 0 \\ 0 & k & -k \end{pmatrix}$, solve for the amounts of each isotope over time.

\item \textbf{Challenge:} Prove that $e^{Jt}$ for a Jordan block $J$ can be computed as:
$e^{Jt} = $e^{\lambda$ t}\sum_{k=0}^{n-1} \frac{t$^{k}$}{k!}N$^{k}$$ where $N = J - \lambda I$.
\end{enumerate}

\section*{Solutions and Hints}

\textbf{Problem 1:} Algebraic multiplicity = 2, geometric multiplicity = 1, defect = 1.

\textbf{Problem 6:} Eigenvector: $\mathbf{v}_{1} = \begin{pmatrix} 1 \\ 0 \end{pmatrix}$, Generalized: $\mathbf{v}_{2} = \begin{pmatrix} 0 \\ 1 \end{pmatrix}$

\textbf{Problem 11:} $\mathbf{x}(t) = $e^{2t}[c_{1}\begin{pmatrix}$ 1 \\ 0 \end{pmatrix} + c_{2}(t\begin{pmatrix} 1 \\ 0 \end{pmatrix} + \begin{pmatrix} 0 \\ 1 \end{pmatrix})]$

\textbf{Problem 16:} Use the standard basis vectors as the Jordan chain.

\textbf{Problem 24:} All solutions decay to zero since $\lambda = -2 < 0$ despite the $t$ and $t^{2}$ terms.

\textbf{Problem 28:} Maximum occurs at $t = 1$ (derivative of $t$e^{-t$}$ equals zero).

\textbf{Key Strategy:} Always check the defect first! If geometric multiplicity equals algebraic multiplicity, use standard eigenvector methods. Otherwise, build Jordan chains systematically.

\textbf{Verification:} For generalized eigenvector $\mathbf{v}_{2}$ satisfying $(A-\lambda I)\mathbf{v}_{2} = \mathbf{v}_{1}$, verify that $\mathbf{x}(t) = $e^{\lambda$ t}(t\mathbf{v}_{1} + \mathbf{v}_{2})$ satisfies $\mathbf{x}' = A\mathbf{x}$.

\end{document}