\documentclass[12pt]{article}
\usepackage{amsmath, amssymb, amsthm, enumitem, geometry}
\geometry{margin=1in}

\title{Lesson 20: Practice Problems - Riccati to Second-Order Transformation}
\author{ODE 1 - Prof. Adi Ditkowski}
\date{}

\usepackage{mdframed}
\usepackage{xcolor}
\usepackage{tikz}
\begin{document}
\maketitle

\section*{Part A: Basic Transformations (5 problems)}

\begin{enumerate}
    \item Transform each Riccati equation to second-order linear form:
    \begin{enumerate}[label=(\alph*)]
        \item $y' = 1 + y^{2}$
        \item $y' = x^{2} - y^{2}$
        \item $y' = \frac{1}{x^{2}} + \frac{2y}{x} + y^{2}$
        \item $y' = e$^{x}$ + y^{2}$
        \item $y' = 1 - 2y + y^{2}$
    \end{enumerate}

    \item For the equation $y' = q_{0}(x) + y^{2}$, show that the transformation $y = -u'/u$ gives $u'' = q_{0}(x)u$.

    \item Verify that if $u_{1}$ and $u_{2}$ solve the second-order equation, then $y = -\frac{u_{1}' u_{2} - u_{1} u_{2}'}{u_{1} u_{2}}$ solves the associated Riccati.

    \item Show that the transformation $y = \tan(\theta)$ converts $y' = a(x)(1 + y^{2})$ into $\theta' = a(x)$.

    \item Prove that if the Riccati has constant coefficients, the transformed second-order equation also has constant coefficients.
\end{enumerate}

\section*{Part B: Complete Solutions via Transformation (6 problems)}

\begin{enumerate}[resume]
    \item Solve $y' = 1 - y^{2}$ by transforming to second-order form.

    \item Find the general solution of $y' = \frac{2}{x^{2}} + y^{2}$ using the second-order approach.

    \item Solve $y' = 4 + y^{2}$ and express the answer in terms of hyperbolic functions.

    \item Transform and solve: $y' = \cos^{2}(x) + y^{2}$.

    \item Find all solutions of $y' = $e^{2x}$ - y^{2}$.

    \item Solve the equation $y' = \frac{1 - n^{2}}{x^{2}} + y^{2}$ where $n$ is an integer.
\end{enumerate}

\section*{Part C: Reverse Transformation (5 problems)}

\begin{enumerate}[resume]
    \item Given $u'' + u = 0$, find the associated Riccati equation and solve it.

    \item Transform $u'' - 4u = 0$ to Riccati form and find all solutions.

    \item Convert the Airy equation $u'' - xu = 0$ to its Riccati form.

    \item Show that the Bessel equation $x$^{2u}$'' + xu' + (x^{2} - n^{2})u = 0$ corresponds to a specific Riccati equation.

    \item Given $u'' + p(x)u' + q(x)u = 0$ with known solution $u_{1}$, find the Riccati solution.
\end{enumerate}

\section*{Part D: Special Cases and Applications (5 problems)}

\begin{enumerate}[resume]
    \item The Riccati equation $y' = ax^{2n} + by^{2}$ where $a, b$ are constants:
    \begin{enumerate}[label=(\alph*)]
        \item Transform to second-order form
        \item Identify when elementary solutions exist
        \item Solve for $n = 0, 1$
    \end{enumerate}

    \item Consider $y' = \frac{A}{x^{2}} + \frac{B}{x}y + Cy^{2}$:
    \begin{enumerate}[label=(\alph*)]
        \item Show this transforms to an Euler equation
        \item Find conditions on $A, B, C$ for real solutions
        \item Solve when $B = 0$, $C = 1$
    \end{enumerate}

    \item The equation $y' = \sec^{2}(x) + y^{2}$:
    \begin{enumerate}[label=(\alph*)]
        \item Transform to second-order form
        \item Explain why elementary solutions don't exist
        \item Find series solution near $x = 0$
    \end{enumerate}

    \item For the parametric family $y' = \lambda + y^{2}$:
    \begin{enumerate}[label=(\alph*)]
        \item Find the second-order form for each $\lambda$
        \item Determine solution behavior as $\lambda$ varies
        \item Identify bifurcation at $\lambda = 0$
    \end{enumerate}

    \item The Schwarzian derivative connection:
    \begin{enumerate}[label=(\alph*)]
        \item Show that $y' = -\frac{1}{2}S[f](x) + y^{2}$ where $S[f]$ is the Schwarzian
        \item Find the second-order form
        \item Discuss invariance properties
    \end{enumerate}
\end{enumerate}

\section*{Part E: Theoretical Problems (4 problems)}

\begin{enumerate}[resume]
    \item Prove that the transformation $y = -\frac{1}{q_{2}}\frac{u'}{u}$ is invertible: given $y(x)$, we can recover $u(x)$ up to a constant multiple.

    \item Show that if the Riccati equation has $n$ particular solutions $y_{1}, \ldots, y_{n}$, the second-order equation has $n$ corresponding solutions $u$_{i}$$ with $y$_{i}$ = -u$_{i}$'/u$_{i}$$.

    \item Prove that the Wronskian $W(u_{1}, u_{2}) = u$_{1u}$_2' - u_{1}'u_{2}$ of two solutions of the second-order equation satisfies $W' = -p(x)W$.

    \item Establish the connection: If $y_{1}$ and $y_{2}$ are two Riccati solutions, then $(y_{1} - y_{2})^{-1}$ satisfies a first-order linear equation.
\end{enumerate}

\section*{Part F: Exam-Style Complete Problems (6 problems)}

\begin{enumerate}[resume]
    \item \textbf{[Prof. Ditkowski Style]}
    Consider the Riccati equation: $y' = \frac{4}{x^{2}} - y^{2}$
    \begin{enumerate}[label=(\alph*)]
        \item Transform to second-order linear form using $y = u'/u$
        \item Identify the type of second-order equation obtained
        \item Solve the second-order equation
        \item Find the general solution of the original Riccati
        \item Verify your solution satisfies the original equation
        \item Find the solution with $y(1) = 2$
    \end{enumerate}

    \item \textbf{[Multiple Methods]}
    For $y' = 1 + y^{2}$:
    \begin{enumerate}[label=(\alph*)]
        \item Solve using the known particular solution $y_{p} = \tan(x)$
        \item Solve by transforming to second-order form
        \item Verify both methods give the same general solution
        \item Discuss the solution's periodicity and singularities
    \end{enumerate}

    \item \textbf{[Comparison Problem]}
    Given the two equations:
    \begin{itemize}
        \item[(i)] $y' = 1 + y^{2}$
        \item[(ii)] $y' = 1 - y^{2}$
    \end{itemize}
    \begin{enumerate}[label=(\alph*)]
        \item Transform both to second-order form
        \item Solve both completely
        \item Compare the qualitative behavior of solutions
        \item Explain the difference using phase portraits
    \end{enumerate}

    \item \textbf{[Application to Quantum Mechanics]}
    The radial Schrödinger equation can yield the Riccati:
    $$y' = \frac{l(l+1)}{x^{2}} - k^{2} + \frac{2m}{\hbar^{2}}V(x) + y^{2}$$
    \begin{enumerate}[label=(\alph*)]
        \item For $V(x) = 0$ (free particle), transform to second-order
        \item Solve for $l = 0$
        \item Discuss bound states vs scattering states
    \end{enumerate}

    \item \textbf{[Comprehensive Problem]}
    Consider $y' = x^{2} + y^{2}$:
    \begin{enumerate}[label=(\alph*)]
        \item Show no elementary particular solution exists
        \item Transform to second-order form
        \item Identify this as an Airy-type equation
        \item Write the solution in terms of Airy functions
        \item Analyze asymptotic behavior as $x \to \pm\infty$
    \end{enumerate}

    \item \textbf{[Challenge: Connection to Painlevé]}
    The equation $y' = x + y^{2}$ is related to Painlevé II.
    \begin{enumerate}[label=(\alph*)]
        \item Transform to second-order form
        \item Show the second-order equation has no elementary solutions
        \item Prove solutions exist for all $x$
        \item Find the asymptotic behavior as $x \to -\infty$
        \item Explain why this is called a "Painlevé transcendent"
    \end{enumerate}
\end{enumerate}

\section*{Solutions and Hints}

\textbf{Selected Solutions:}

\textbf{Problem 1(a):} Using $y = -u'/u$ with $q_{2} = 1$:
Second-order form: $u'' + u = 0$
Solution: $u = c_{1}\cos x + c_{2}\sin x$
Riccati solution: $y = \tan(x - \phi)$

\textbf{Problem 6:} $y' = 1 - y^{2}$ transforms to $u'' - u = 0$.
Solution: $u = c$_{1e}$^x + c$_{2e}$^{-x}$
Therefore: $y = -\frac{c$_{1e}$^x - c$_{2e}$^{-x}}{c$_{1e}$^x + c$_{2e}$^{-x}} = \tanh(x + C)$

\textbf{Problem 7:} $y' = \frac{2}{x^{2}} + y^{2}$ gives $x$^{2u}$'' + 2u = 0$.
This is Euler with $m(m-1) + 2 = 0$, so $m = \frac{1 \pm \sqrt{1-8}}{2} = \frac{1 \pm i\sqrt{7}}{2}$.
Solutions involve $x^{1/2}\cos(\frac{\sqrt{7}}{2}\ln x)$ and $x^{1/2}\sin(\frac{\sqrt{7}}{2}\ln x)$.

\textbf{Problem 12:} From $u'' + u = 0$, the Riccati is $y' = -1 - y^{2}$.
This is $y' = 1 + y^{2}$ with $y \to iy$, giving $y = -\tan(x - C)$.

\textbf{Problem 25:} For $y' = \frac{4}{x^{2}} - y^{2}$:
Second-order: $x$^{2u}$'' - 4u = 0$ (Euler equation)
With $u = x$^{m}$$: $m(m-1) = 4$, so $m = \frac{1 \pm \sqrt{1+16}}{2} = \frac{1 \pm \sqrt{17}}{2}$

\textbf{Key Transformation Formulas:}
\begin{itemize}
    \item Forward: $y = -\frac{1}{q_{2}} \cdot \frac{u'}{u}$
    \item Resulting 2nd-order: $u'' + p(x)u' + r(x)u = 0$
    \item Where: $p = -q_{1} - q_{2}'/q_{2}$, $r = q$_{0q}$_2$
    \item Reverse: Given $u'' + p(x)u' + q(x)u = 0$, get $y' = -q - py - y^{2}$
\end{itemize}

\textbf{Common Second-Order Results:}
\begin{itemize}
    \item $y' = a + y^{2} \to u'' + au = 0$
    \item $y' = a - y^{2} \to u'' - au = 0$
    \item $y' = \frac{a}{x^{2}} + y^{2} \to x$^{2u}$'' + au = 0$
    \item $y' = ax$^{n}$ + y^{2} \to u'' + ax$^{nu}$ = 0$
\end{itemize}

\end{document}