\documentclass[12pt]{article}
\usepackage{amsmath, amssymb, amsthm, enumitem, geometry}
\geometry{margin=1in}

\title{Lesson 19: Practice Problems - Riccati Equations with Known Solution}
\author{ODE 1 - Prof. Adi Ditkowski}
\date{}

\usepackage{mdframed}
\usepackage{xcolor}
\usepackage{tikz}
\begin{document}
\maketitle

\section*{Part A: Recognition and Classification (5 problems)}

\begin{enumerate}
    \item Identify which equations are Riccati and find $q_0$, $q_1$, $q_2$:
    \begin{enumerate}[label=(\alph*)]
        \item $y' = x^2 + 2xy - y^2$
        \item $y' = $e^x$ + y^2$
        \item $xy' = 1 + xy + y^2$
        \item $y' + y = xy^2$
        \item $y' = \sin x + 2y\cos x + y^2$
    \end{enumerate}

    \item Show that the equation $y' = \frac{a + by}{c + dy}$ can be written in Riccati form if and only if $ad - bc \neq 0$.

    \item Verify that if $y_1$ and $y_2$ are two solutions of a Riccati equation, then $y = y_1 + \frac{1}{z}$ satisfies a linear equation in $z$.

    \item Prove that the sum of two particular solutions of a Riccati equation does not generally give another solution.

    \item For the Riccati equation $y' = q_0 + q_1y + q_2y^2$, show that $y = -q_1/(2q_2)$ is a solution if and only if $q_0 = q_1^2/(4q_2)$.
\end{enumerate}

\section*{Part B: Finding Particular Solutions (6 problems)}

\begin{enumerate}[resume]
    \item Find a particular solution by inspection:
    \begin{enumerate}[label=(\alph*)]
        \item $y' = 2 + y - y^2$
        \item $y' = \frac{1}{x^2} - \frac{y}{x} + y^2$
        \item $y' = 1 + 2y + y^2$
        \item $y' = $e^{2x}$ + $e^x$ y - y^2$
    \end{enumerate}

    \item Verify that $y_p = \tan x$ satisfies $y' = 1 + y^2$ and find the general solution.

    \item Given that $y_p = 1/x$ is a solution of $y' = -1/x^2 + 2y/x - y^2$, find all solutions.

    \item Show that $y_p = x$ satisfies $y' = 1 - x^2 + 2xy - y^2$ and solve completely.

    \item Find two different particular solutions of $y' = 6 - y - y^2$ and use each to find the general solution.

    \item For $y' = 2\cos^2 x + (\sin 2x)y - y^2$, verify that $y_p = \sin x$ is a solution.
\end{enumerate}

\section*{Part C: Complete Solution Process (5 problems)}

\begin{enumerate}[resume]
    \item Solve the Riccati equation $y' = \frac{2}{x^2} - \frac{2y}{x} + y^2$ given that $y_p = 2/x$.

    \item Find all solutions of $y' = -2 + y + y^2$ given one solution $y_p = 1$.

    \item Solve $y' = $e^{2x}$ + (1 - 2$e^x)y$ + y^2$ with the particular solution $y_p = e^x$.

    \item Given $y_p = \cot x$ solves $y' = -1 - y^2$, find the solution satisfying $y(\pi/4) = 0$.

    \item Solve $y' = \frac{1 - x^2}{x^2} + \frac{2y}{x} - y^2$ knowing that it has a polynomial particular solution.
\end{enumerate}

\section*{Part D: Advanced Problems (5 problems)}

\begin{enumerate}[resume]
    \item Consider the parametric family $y' = a + y^2$ where $a$ is a constant.
    \begin{enumerate}[label=(\alph*)]
        \item Find particular solutions for $a = 1, 0, -1$
        \item Solve each case completely
        \item Discuss the qualitative behavior of solutions
    \end{enumerate}

    \item The equation $y' = q(x) + y^2$ where $q(x)$ is continuous:
    \begin{enumerate}[label=(\alph*)]
        \item Show that if $q(x) = -f'(x)/f(x)$ for some $f(x) > 0$, then $y_p = f'(x)/(2f(x))$ is a solution
        \item Apply this to $q(x) = -2x/(1 + x^2)$
    \end{enumerate}

    \item Solve the Riccati equation arising in optimal control:
    $$y' = 1 - y^2$$
    with $y(0) = 0$.

    \item For the equation $y' = x^{2n} + y^2$ where $n$ is a positive integer:
    \begin{enumerate}[label=(\alph*)]
        \item Show there's no polynomial particular solution
        \item Transform to a second-order linear equation
        \item Find the solution for $n = 0$
    \end{enumerate}

    \item Consider the Riccati equation with periodic coefficients:
    $$y' = \cos(2x) + 2\sin x \cdot y - y^2$$
    Given $y_p = \sin x$, find all periodic solutions.
\end{enumerate}

\section*{Part E: Theoretical Problems (4 problems)}

\begin{enumerate}[resume]
    \item Prove that if a Riccati equation has three known particular solutions $y_1, y_2, y_3$, then the general solution can be written as:
    $$\frac{y - y_1}{y - y_2} = C \cdot \frac{y_3 - y_1}{y_3 - y_2}$$

    \item Show that the transformation $y = -u'/u$ converts the second-order linear equation $u'' + p(x)u' + q(x)u = 0$ into the Riccati equation:
    $$y' = -q(x) - p(x)y - y^2$$

    \item Prove that if $y_1$ and $y_2$ are two solutions of a Riccati equation, then:
    $$\frac{d}{dx}\left(\frac{1}{y_1 - y_2}\right) = -q_1 - q_2(y_1 + y_2)$$

    \item For the autonomous Riccati $y' = a + by + cy^2$:
    \begin{enumerate}[label=(\alph*)]
        \item Find conditions for existence of equilibrium points
        \item Analyze stability of equilibria
        \item Show that solutions either blow up in finite time or exist for all time
    \end{enumerate}
\end{enumerate}

\section*{Part F: Exam-Style Complete Problems (5 problems)}

\begin{enumerate}[resume]
    \item \textbf{[Prof. Ditkowski Style]}
    Consider the equation: $y' = \frac{4}{x^2} - \frac{4y}{x} + y^2$
    \begin{enumerate}[label=(\alph*)]
        \item Verify that $y_p = 2/x$ is a particular solution
        \item Use the substitution $y = y_p + v$ to transform to Bernoulli form
        \item Solve the resulting Bernoulli equation
        \item Find the general solution
        \item Determine the solution satisfying $y(1) = 3$
        \item Are there any singular solutions?
    \end{enumerate}

    \item \textbf{[Comprehensive Problem]}
    For the equation $y' = 1 + xy - y^2$:
    \begin{enumerate}[label=(\alph*)]
        \item Show that no constant particular solution exists
        \item Try $y_p = ax + b$ and find values of $a$ and $b$
        \item Solve the equation completely
        \item Analyze behavior as $x \to \pm\infty$
    \end{enumerate}

    \item \textbf{[Multiple Methods]}
    Given $y' = 2 - 3y + y^2$:
    \begin{enumerate}[label=(\alph*)]
        \item Find two different particular solutions
        \item Use each to find the general solution
        \item Verify both give the same general solution
        \item Express the solution using partial fractions
    \end{enumerate}

    \item \textbf{[Application to Projectile Motion]}
    The equation for the envelope of projectile trajectories:
    $$y' = \frac{g}{2v_0^2}x + \sqrt{1 + \left(\frac{gx}{v_0^2}\right)^2} - \frac{g^2x}{2v_0^4}y^2$$
    \begin{enumerate}[label=(\alph*)]
        \item Show this is approximately Riccati for small $x$
        \item Find the linear approximation
        \item Discuss physical interpretation
    \end{enumerate}

    \item \textbf{[Challenge Problem]}
    Consider the family of Riccati equations:
    $$y' = \frac{n(n+1)}{x^2} - \frac{2n}{x}y + y^2$$
    where $n$ is a positive integer.
    \begin{enumerate}[label=(\alph*)]
        \item Show that $y_p = n/x$ is always a particular solution
        \item Find the general solution for arbitrary $n$
        \item What happens as $n \to \infty$?
        \item Connect to Legendre polynomials
    \end{enumerate}
\end{enumerate}

\section*{Solutions and Hints}

\textbf{Selected Solutions:}

\textbf{Problem 1(a):} Not Riccati (wrong sign on $y^2$ term). Would need $+y^2$.

\textbf{Problem 6(a):} Try constants: $0 = 2 + c - c^2$, so $c^2 - c - 2 = 0$. Thus $c = 2$ or $c = -1$.

\textbf{Problem 7:} With $y = \tan x + v$: $v' = 2\tan x \cdot v + v^2$.
Using $w = 1/v$: $w' = -2\tan x \cdot w - 1$.
Solution: $w = (\cos x)(C - x)$, so $y = \tan x + \frac{1}{(\cos x)(C - x)}$.

\textbf{Problem 12:} With $y = 2/x + v$:
$v' = \frac{2v}{x} + v^2$ (Bernoulli with $n = 2$).
Let $w = 1/v$: $w' = -\frac{2w}{x} - 1$.
Solution: $w = \frac{C}{x^2} - \frac{x}{3}$.

\textbf{Problem 25:} For equilibria: $0 = a + by + cy^2$.
Discriminant $\Delta = b^2 - 4ac$ determines number of equilibria.
Stability: Check $f'(y^*) = b + 2cy^*$.

\textbf{Key Transformation Formulas:}
\begin{itemize}
    \item Riccati: $y' = q_0 + q_1y + q_2y^2$
    \item If $y_p$ known: $y = y_p + v$
    \item Bernoulli form: $v' = (q_1 + 2q_2y_p)v + q_2v^2$
    \item Linear form: $w' = -(q_1 + 2q_2y_p)w - q_2$ where $w = 1/v$
\end{itemize}

\textbf{Common Particular Solutions:}
\begin{itemize}
    \item Constants when $q_0, q_1, q_2$ are constants
    \item $y = a/x$ for equations with $1/x^2$ terms
    \item $y = \tan(ax)$ for $y' = a^2 + y^2$
    \item $y = \tanh(ax)$ for $y' = -a^2 + y^2$
\end{itemize}

\end{document}